Inhibitory basal ganglia nuclei differentially innervate pedunculopontine nucleus subpopulations and evoke opposite motor and valence behaviors

Version 1

Abstract

Summary

The canonical basal ganglia model predicts that the substantia nigra pars reticulata (SNr) and the globus pallidus externa (GPe) will have specific effects on locomotion: the SNr inhibiting locomotion and the GPe enhancing it. In this manuscript, we use in vivo optogenetics to show that a projection-defined neural subpopulation within each structure exerts non-canonical effects on locomotion. These non-canonical subpopulations are defined by their projection to the pedunculopontine nucleus (PPN) and mediate opposing effects on reward. To understand how these structures differentially modulate the PPN, we use ex vivo whole-cell recording with optogenetics to comprehensively dissect the SNr and GPe connections to regionally– and molecularly-defined populations of PPN neurons. The SNr inhibits all PPN subtypes, but most strongly inhibits caudal glutamatergic neurons. The GPe selectively inhibits caudal glutamatergic and GABAergic neurons, avoiding both cholinergic and rostral cells. This circuit characterization reveals non-canonical basal ganglia pathways for locomotion and valence.

Introduction

The pedunculopontine nucleus (PPN) is a brainstem structure heavily interconnected with subcortical structures, such as the basal ganglia, thalamus, and spinal cord. The PPN has been implicated in locomotor control and valence processing1–4. However, understanding PPN circuitry is complicated by its anatomical and molecular heterogeneity. It has three molecularly-defined cell types, distinguished by their major neurotransmitter5 and distinct rostral and caudal topography. While cholinergic and glutamatergic neurons are more densely packed in the caudal PPN, GABAergic neurons are more densely packed in the rostral PPN5,6. These molecularly-defined populations display very minimal overlap5,7,8. This unique topography has been conserved across species9–11, but it is unknown how the basal ganglia connects to each of the distinct cell types in the rostral and caudal PPN regions to modulate locomotor and valence outputs.

While the PPN interacts with all basal ganglia nuclei3,9, the inhibitory inputs from the substantia nigra pars reticulata (SNr) and the globus pallidus externus (GPe) to the PPN are of particular interest because of their roles in the canonical motor pathways for ‘stop’ and ‘go,’ respectively. Both the SNr and GPe are known to send axonal projections to the PPN12–18. Previous studies have identified SNr and GPe monosynaptic connections to the PPN cholinergic and glutamatergic neurons using rabies tracing13–17. However, inputs to the GABAergic neurons have only been described in terms of the entire mesencephalic locomotor region (MLR), which extends beyond the PPN14. While SNr projections to MLR GABAergic neurons were identified but cannot be isolated to the PPN, GPe projections to the GABAergic MLR neurons were not detected14. Of note, these previous studies consider the PPN as a whole rather than separating its rostral and caudal subregions. While viral tracing can provide insight into neuroanatomical connections, limitations of these techniques19–21 and evidence that the rostral and caudal PPN may have distinct functions22–24 prompt a need for electrophysiological studies to determine whether the SNr and GPe inputs differentially innervate regionally-defined PPN neurons.

Optogenetic manipulations of the molecularly-defined PPN cell types have resulted in variable, and sometimes contradictory, locomotor and valence behaviors7,14,16,17,22–32. More recent studies have refined our understanding of these opposing behavioral effects by selectively stimulating the rostral or caudal GABAergic and glutamatergic PPN neurons22–24. In the basal ganglia, direct stimulation of molecularly-defined SNr and GPe subpopulations can also have distinct effects on locomotion and valence33–41. Despite the potential for the basal ganglia to modify behavior through the motor brainstem14, it is not clear how the SNr and GPe influence the rostral and caudal PPN to modify locomotion and valence.

Here, we use whole-cell patch clamp electrophysiology paired with optogenetics to selectively stimulate the SNr or GPe axons while recording inhibitory inputs to the cholinergic, GABAergic, and glutamatergic PPN neurons to comprehensively characterize the strength and pattern of inhibition across the rostrocaudal PPN axis. We find that the GPe preferentially inhibits caudal GABAergic and glutamatergic PPN neurons whereas the SNr most strongly inhibits a caudal ‘hotspot’ of PPN glutamatergic neurons. Stimulating SNr or GPe axons over the PPN in vivo evokes opposing valence processing outcomes with place aversion during SNr stimulation and place preference during GPe stimulation. Surprisingly, and counter to the predictions of the canonical basal ganglia model, we find that stimulating PPN-projecting SNr neurons increases locomotion, while the GPe decreases locomotion through its projections to the PPN. While both the SNr and GPe inhibit the PPN, our results show that each nucleus differentially modulates the activity of specific cell types in the rostral and caudal PPN and are implicated in non-canonical basal ganglia circuits for modulating locomotion and valence processing.

Results

SNr and GPe axons display distinct distribution patterns across the rostral and caudal PPN

To characterize the axon distribution pattern from the SNr and GPe throughout the rostral and caudal extent of the PPN, we stereotaxically injected a channelrhodopsin virus with a fluorescent marker (AAV1-hSyn-ChR2-eYFP) into either the SNr or GPe of ChAT-Cre/Ai9-tdTomato mice (Figure 1A&C). We sliced 200 μm sagittal slices from mouse brains three weeks after viral injection and cleared the brain slices using the CUBIC clearing method42. Using the red fluorescent ChAT-positive neurons to define the borders of the PPN, we found that SNr axons appear in both the rostral and caudal PPN. The SNr axons fill the rostral PPN evenly, but appear more distinctly clustered in specific areas of the caudal PPN (Figure 1B), whereas GPe axonal projections more densely fill the caudal region and strikingly avoid the rostral region (Figure 1D). These anatomical findings predict that the SNr will inhibit both rostral and caudal PPN neurons, while the GPe will selectively inhibit the caudal PPN. However, this axon imaging method does not reveal whether these axons form functional synapses with the neurons in the PPN and does not provide information on the strength of inhibition from each structure onto specific PPN cell types. Therefore, to evaluate the functional connectivity and synaptic characteristics from these inhibitory basal ganglia output structures to regionally– and molecularly-defined PPN neural subpopulations, we systematically recorded the inhibitory input to the rostral and caudal PPN from the SNr and GPe in three genetic mouse lines (ChAT-cre, Vgat-cre, and Vglut2-cre) to identify the cholinergic, GABAergic, and glutamatergic PPN subpopulations, respectively (see Figure S1).

SNr and GPe axons display distinct distribution patterns across the rostral and caudal PPN.

(A,C) Stereotaxic injection of AAV1 delivering hSyn-ChR2-eYFP to the SNr or GPe of ChAT-Cre/Ai9-tdTomato mice, respectively. (B, D) Confocal images of EYFP-filled SNr or GPe axons across the PPN, respectively. CNII: cranial nerve II, scp: superior cerebellar peduncle; PB: parabrachial nucleus

The SNr inhibits cholinergic neurons in both the rostral and caudal PPN to a similar extent, but evokes stronger rebound in rostral neurons

To evaluate the functional connection between the SNr and cholinergic PPN neurons, we performed ex vivo whole-cell patch clamp electrophysiology paired with optogenetics to record synaptic currents in ChAT-Cre/Ai9-tdTomato mice injected with ChR2 in the SNr three weeks prior (Figure 2A). Neurons were targeted for whole-cell patch clamp in both the rostral and caudal PPN across three sagittal slices to evaluate the medial to lateral extent of the PPN (Figure 2B). To measure optically-evoked inhibitory post-synaptic currents (oIPSCs), cholinergic PPN neurons were held at –50 mV in whole-cell voltage clamp and blue (470 nm) light was applied (2 ms pulse duration, 13 mW) to activate the ChR2-infected SNr axons. In connected neurons, inhibitory currents were recorded in the presence of glutamatergic receptor blockers (AP5, NBQX, CNQX) during stimulation (Figure 2C). In a subset of neurons, a GABA-a receptor blocker, GABAzine, was applied to confirm the oIPSCs were GABA-mediated (Figure 2C). GABAzine eliminated SNr-evoked inhibitory currents in most cells, but a small residual current remained in some cells (see Methods for details). We found that all rostral (n=19/19 cells, N=6 mice) and essentially all caudal (n=26/27, N=6) recorded cholinergic PPN neurons responded to SNr axon activation with an observable oIPSC (Figure 2D).

SNr inhibition of rostral and caudal ChAT+ PPN neurons.

(A) Experimental set up to identify red ChAT+ PPN neurons for whole-cell patch clamp while stimulating ChR2-filled SNr axons [N=6]. (B) White arrowheads pointing to neurobiotin-filled patched neurons within the PPN across three 200µm slices. (C) Example trace of the first five oIPSCs [blue] in the 2-second 20 Hz train inhibited by GABA-a receptor blocker, GABAzine [green], while holding the cell at –50mV. (D) Percent connected among patched neurons in the rostral and caudal regions. (E) Average oIPSC amplitude at each of 40 optogenetic light pulses in n=15 rostral neurons and n=20 caudal neurons. (F) Left, Individual cell data for the first oIPSC amplitude and, right, example current traces. (G) Cell mapping of patched neuron locations with the first oIPSC amplitude represented by the color scale. (H) Normalized current amplitudes in E. (I) Left, Individual cell data for the PPR between the first two oIPSC amplitudes in the train and, right, example current traces. (J) Example voltage traces of action potential firing during a 2-second 20 Hz train stimulation in rostral (left) and caudal (right) neurons. (K) Percent of pre-optical stimulation firing frequency during stimulation and rebound in n=14 rostral vs n=23 caudal neurons; rebound rostral vs caudal p=0.0310. (L) Individual cell data for the absolute change in frequency during optical stimulation [ΔFrq During Opto]. (M) Individual cell data for the absolute change in rebound frequency post-stimulation [ΔRebFrq]; rostral vs. caudal p=0.0142. (N) Correlation analysis, color scale representing Spearman r [-1,1] and size representing p-value [1,0]. (O) Negative correlation between the absolute change in frequency during stimulation and post-stimulation rebound; r=-0.372, p=0.039. * p<0.05; bar graph data represent mean ± SEM; box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles.

To determine whether the SNr input to the rostral and caudal PPN cholinergic neurons differed in synaptic strength, we measured the amplitude of each oIPSC in a 20 Hz train of 2 ms optical stimulations applied for 2 seconds. We found that the oIPSCs ranged from tens to hundreds of picoamperes (pA) in both groups and there was no significant difference between the average amplitude of the first oIPSC in the stimulus train onto rostral vs caudal cholinergic neurons (mean ± SEM; first oIPSC n=15 Rostral: 92.6 ± 18 pA, n=20 Caudal: 115.2 ± 17 pA; unpaired t-test, t=0.9041, p=0.3725; Figure 2E&F). Going forward, the first oIPSC amplitude of the train will be referred to as the first oIPSC.

While the rostral and caudal anatomical separation of PPN neurons is commonly used, it remains a relatively coarse division. Therefore, we carefully mapped the location of each recorded cholinergic PPN neuron by matching slices to their bregma reference in the Paxinos and Franklin’s mouse brain atlas and aligning the midline of cholinergic neuron distribution which separates the loosely spread rostral neurons and densely packed caudal neurons (See Methods for details, Figure 2G). The strength of the SNr inhibitory connection is depicted by a color scale representative of the first oIPSC (Figure 2G, right). Throughout both the rostral and caudal PPN, neurons receiving larger inputs from the SNr are intermixed with neurons receiving smaller inputs.

The postsynaptic response to a 2-second 20 Hz stimulation train can undergo short-term synaptic plasticity which either depresses or facilitates current amplitude with subsequent stimulations. To compare short-term synaptic plasticity characteristics between rostral and caudal cholinergic PPN neurons, we normalized the amplitude of each current in the train to the first current. We found that the oIPSC amplitude remained relatively constant with each subsequent pulse in both rostral and caudal cholinergic neurons (Figure 2H). In addition, the paired-pulse ratio (PPR) of the peak of the first two currents in the stimulation train did not differ (mean ± SEM; PPR n=15 Rostral: 0.96 ± 0.04, n=20 Caudal: 0.97 ± 0.05; unpaired t-test, t=0.1082, p=0.9145; Figure 2I). While the postsynaptic currents in some cells displayed short-term synaptic plasticity, these data show that SNr synaptic inputs onto cholinergic neurons do not undergo significant short-term synaptic plasticity in either the rostral or caudal PPN.

Although oIPSC amplitude is indicative of the strength of a synapse, it does not always correlate with the functional impact on a neuron’s action potential output20,43,44. Because cholinergic PPN neurons display spontaneous firing45–49, we characterized the impact of SNr-mediated inhibition on action potential firing rate. While optically stimulating SNr axons with the same 20 Hz blue light protocol as above, we observed a decrease in firing rate in the recorded cholinergic PPN neuron (Figure 2J). Both rostral and caudal neuronal populations showed significant inhibition compared to their pre-optical stimulation firing (2way ANOVA, p=0.0154; Tukey test, %Frequency Pre vs During Stimulation Rostral p<0.0001, Pre vs During Caudal p<0.0001), and there was no significant difference between the percent frequency change in rostral and caudal neurons during SNr axon stimulation (mean ± SEM; %Frequency During Stimulation n=14 Rostral: 35.62 ± 8.87 %, n=23 Caudal: 52.11 ± 8.43 %; 2way ANOVA, p=0.0154; Tukey test, Rostral vs Caudal p=0.1133; Figure 2K). There was also no rostrocaudal difference in cholinergic neuron firing when measuring the absolute decrease in firing frequency during optical stimulation [median (IQR); ΔFrq During Opto n=14 Rostral: – 2.47 Hz (–3.37 to –1.44 Hz), n=23 Caudal: –2.10 Hz (–3.11 to –0.84 Hz); Mann Whitney, U=131, p=0.3602; Figure 2L]. These data show that the SNr equally inhibits action potential firing in rostral and caudal PPN cholinergic neurons.

Post-inhibitory rebound firing has been observed in the PPN as well as other brainstem structures48,50–55 and is associated with greater excitability, temporal encoding, and generation of oscillatory activity56. Indeed, we found that the rostral and caudal cholinergic PPN neurons exhibited an increase in action potential firing after inhibition (2way ANOVA, p=0.0154; Tukey test, %Frequency Pre vs. Post Stimulation Rostral p=0.0013, Pre vs Post Caudal p<0.0012; Figure 2K). This is in agreement with previous literature showing rebound spikes following inhibitory postsynaptic potentials recorded in PPN neurons of male rats during electrical stimulation of the SNr50,51. However, we found that the rostral cholinergic neurons experienced greater increases in post-inhibitory rebound excitation compared with caudal cholinergic neurons (mean ± SEM; %Frequency Post Stimulation n=14 Rostral: 136.1 ± 7.8 %, n=23 Caudal: 119.1 ± 4.6 %; 2way ANOVA, p=0.0154; Tukey test, Rostral vs Caudal p=0.0310; Figure 2K). Rostral neurons also show a significantly larger increase in rebound activity as measured by the absolute change in rebound frequency [median (IQR); ΔRebFrq n=14 Rostral: 1.64 Hz (1.00 to 2.17 Hz), n=22 Caudal: 0.83 Hz (0.42 to 1.14 Hz); Mann Whitney, U=79, p=0.0142; Figure 2M]. This aligns with previous findings showing that rostral cholinergic neurons tend to have stronger intrinsic rebound activity compared to caudal neurons48. Importantly, we show that such rebound can be evoked by synaptic inhibition from the SNr.

To detect trends among SNr connectivity characteristics and intrinsic electrophysiological properties of cholinergic PPN neurons, we performed a correlation analysis to determine the strength and directionality of the relationship among five parameters [first oIPSC, PPR, ΔFrq During Opto, pre-optical stimulation frequency (Pre-Opto Frq), and ΔRebFrq] (Figure 2N). We found a weak but significant negative correlation between the frequency decrease during SNr axon stimulation and the rebound frequency increase (Spearman r=-0.372, p=0.039, Figure 2N&O). This finding is expected since strong inhibition is more effective at producing rebound activity than weak inhibition. However, rostral PPN cholinergic neurons do not receive significantly stronger inhibition from the SNr compared with caudal cholinergic neurons (Figure 2F&L). Therefore, larger rebound frequency in rostral neurons is likely due to differences in intrinsic cellular properties48. Put together, these findings show that the SNr inhibits rostral and caudal cholinergic PPN neurons to a similar extent, but evokes stronger rebound firing in the rostral neurons.

The SNr inhibits GABAergic neurons in both the rostral and caudal PPN to a similar extent

While there are several rabies tracing studies suggesting SNr axons project to the cholinergic and glutamatergic PPN neurons13–17, SNr projections to the GABAergic PPN neurons are inferred from a study of the entire mesencephalic locomotor region14 and a characterization of their synaptic strength has not been conducted. Therefore, we evaluated the synaptic characteristics between the SNr and GABAergic neurons in the rostral and caudal PPN. In Vgat-Cre/Ai9-tdTomato mice, we injected ChR2 into the SNr to optically stimulate axons over the PPN while patching Vgat+ GABAergic PPN neurons (Figure 3A). We post-hoc stained the slices for ChAT+ neurons to verify the location of all patched GABAergic neurons within the cholinergic PPN. We found that all rostral (n=12/12, N=6) and most caudal (n=20/22, N=6) recorded GABAergic neurons received inhibitory input from the SNr (Figure 3B).

SNr inhibition of rostral and caudal Vgat+ PPN neurons.

(A) Experimental set up to identify red Vgat+ PPN neurons for whole-cell patch clamp while stimulating ChR2-filled SNr axons [N=6]. (B) Percent connected among patched neurons in the rostral and caudal regions. (C) Average oIPSC amplitude at each of 40 optogenetic light pulses in n=9 rostral neurons and n=13 caudal neurons. (D) Left, Individual cell data for the first oIPSC amplitude and, right, example current traces. (E) Cell mapping of patched neuron locations with the first oIPSC amplitude represented by the color scale. (F) Normalized current amplitudes in C. (G) Left, Individual cell data for the PPR between the first two oIPSC amplitudes in the train and, right, example current traces. (H) Percent of pre-optical stimulation firing frequency [% Pre-Opto Frq] during stimulation and and rebound in n=7 rostral and n=19 caudal neurons. (I) Individual cell data for the absolute change in frequency during optical stimulation [ΔFrq During Opto]. (J) Correlation analysis, color scale representing Spearman r [-1,1] and size representing p-value [1,0]. (K) Negative correlation between the absolute change in frequency during stimulation and first oIPSC amplitude; r=-0.755, p=0.001. (L) Negative correlation between the pre-optical stimulation firing frequency and the PPR; r =-0.706, p=0.002. Box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles.

Although there was no major difference in the proportion of GABAergic neurons receiving inhibitory input from the SNr in the rostral vs caudal PPN, we compared the current amplitudes to characterize the synaptic strength of this connection. Both the oIPSC amplitudes across the 2 second stimulation and the first oIPSC showed no difference between rostral and caudal GABAergic neurons [median (IQR); first oIPSC n=9 Rostral: 76.5 pA (27.0 to 209.9 pA), n=13 Caudal: 59.7 pA (37.7 to 269.7 pA); Mann Whitney, U=50, p=0.6005; Figure 3C&D]. While the first oIPSC was not significantly different between rostral and caudal GABAergic neurons, the variability among caudal GABAergic neurons shows an interesting division between a subgroup responding with very small (<60 pA) amplitude currents and another subgroup responding with currents more than triple that amplitude, with one neuron responding with a 2.2 nA inhibitory current. We mapped the location of each recorded GABAergic neuron, as described above, and found no specific location corresponding with the neurons receiving stronger SNr inhibition across the PPN landscape (Figure 3E). These findings suggest that GABAergic PPN neurons may be even further divided into distinct subpopulations defined by either strong or weak SNr input.

We also did not observe any differences in short-term synaptic plasticity between rostral and caudal GABAergic neurons (Figure 3F) and found no significant difference in PPR between rostral and caudal GABAergic neurons [median (IQR); PPR n=9 Rostral: 1.01 (0.88 to 1.16), n=13 Caudal: 1.06 (0.96 to 1.15); Mann Whitney, U=48, p=0.5123; Figure 3G]. This shows that, similar to the SNr connection with cholinergic PPN neurons, SNr input to the GABAergic PPN does not show significant short-term synaptic plasticity in either the rostral or caudal region.

Because GABAergic PPN neurons display spontaneous firing49, we recorded action potential firing in current clamp while optically stimulating the SNr axons. The percent inhibition was not significantly different between rostral and caudal neurons (mean ± SEM; %Frequency During Stimulation n=7 Rostral: 37.91 ± 15.91 %, n=19 Caudal: 30.68 ± 8.03 %; 2way ANOVA, p=0.4662; Figure 3H), and there was no difference in rebound firing frequency (mean ± SEM; %Frequency Post Stimulation n= 7 Rostral: 100.0 ± 14.6 %, n=19 Caudal: 113.2 ± 5.2 %; 2way ANOVA, p=0.4662; Figure 3H). There was also no difference in the absolute frequency change during optical stimulation between rostral and caudal neurons [median (IQR); ΔFrq During Opto n=7 Rostral: –5.03 (–5.55 to –1.34) Hz, n=19 Caudal: –4.29 (–11.02 to –2.94) Hz; Mann Whitney, U=55, p=0.5336; Figure 3I]. These data show that the SNr equally inhibits action potential firing in rostral and caudal GABAergic PPN neurons and does not evoke significant rebound firing in this neural population.

While we found that the SNr inhibits the rostral and caudal GABAergic PPN neurons to a similar extent, there was greater variability in synaptic strength and impact on neuronal action potential firing among caudal GABAergic neurons. This strong variability suggests that GABAergic PPN neurons may be divided into functionally distinct neural types that do not correspond to their rostrocaudal anatomical location. To determine if the neurons receiving larger inhibitory currents have distinct characteristics, we performed a correlation analysis (Figure 3J). As expected, we found that the absolute frequency reduction during inhibition is correlated with the first oIPSC amplitude recorded in a cell (Spearman r=-0.755, p=0.001; Figure 3K). Surprisingly, however, we found a significant negative correlation between PPR and the pre-optical stimulation frequency (Pre-Opto Frq). Specifically, input to slower-firing neurons displayed greater short-term synaptic facilitation. This finding supports the idea that multiple functional neural subpopulations may be present within the GABAergic PPN population. Together, these results show that the SNr functionally inhibits GABAergic neurons across the rostrocaudal extent of the PPN, but suggests the presence of additional heterogeneity within the GABAergic PPN population.

The SNr differentially inhibits the rostral and caudal glutamatergic PPN neurons

Previous rabies tracing anatomical studies have shown SNr inputs to glutamatergic PPN neurons14,16,17, but a full characterization of their synaptic strength and rostral and caudal connectivity has not been conducted. In Vglut2-Cre/Ai9-tdTomato mice, we injected ChR2 into the SNr to optically stimulate axons over the PPN while patching Vglut2+ glutamatergic neurons in the PPN (Figure 4A). While holding cells at –50mV and applying 470 nm blue light stimulation, neurons were identified as connected if oIPSCs were observed. Similar to our findings in cholinergic and GABAergic neurons, we found that all rostral (n=19/19, N=6) and all caudal (n=28/28, N=6) recorded glutamatergic neurons received inhibitory input from the SNr (Figure 4B).

SNr inhibition of rostral and caudal Vglut2+ PPN neurons.

(A) Experimental set up to identify red Vglut2+ PPN neurons for whole-cell patch clamp while stimulating ChR2-filled SNr axons [N=6]. (B) Percent connected among patched neurons in the rostral and caudal regions. (C) Average oIPSC amplitude at each of 40 optogenetic light pulses in n=13 rostral neurons and n=13 caudal neurons. (D) Left, Individual cell data for the first oIPSC amplitude and, right, example current traces; p=0.0035. (E) Cell mapping of patched locations with the first oIPSC amplitude represented by the color scale. (F) Normalized current amplitudes in C. (G) Left, Individual cell data for the PPR between the first two oIPSC amplitudes in the train and, right, example current traces. (H) Example voltage traces of action potential firing during a 2-second 20 Hz train stimulation in rostral and caudal neurons, top to bottom. (I) Percent of pre-optical stimulation firing frequency [% Pre-Opto Frq] during stimulation and rebound in n=13 rostral and n=17 caudal neurons. (J) Individual cell data for the absolute change in frequency during optical stimulation [ΔFrq During Opto]; p=0.0197. (K) Spontaneous frequency in n=11 rostral and n=16 caudal neurons; p=0.0343. (L) Correlation analysis, color scale representing Spearman r [-1,1] and size representing p-value [1,0]. (M) Positive correlation between the absolute change in frequency during stimulation and PPR, r=0.486, p=0.030. (N) Negative correlation between the absolute change in frequency during stimulation and first oIPSC amplitude; r=-0.841, p<0.00001. (O) Negative correlation between the absolute change in frequency during stimulation and pre-optical stimulation frequency; r=-0.791, p<0.0001. (P) Positive correlation between the first oIPSC amplitude and pre-optical stimulation frequency; r=0.818, p<0.0001. * p<0.05, ** p<0.01; box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles.

Comparing the oIPSC amplitudes across the 20 Hz train, we discovered that the caudal glutamatergic neurons receive larger inhibitory currents than rostral glutamatergic neurons (Figure 4C). The median first oIPSC measured in caudal glutamatergic neurons was also significantly larger than those recorded in rostral neurons [median (IQR); first oIPSC n=13 Rostral: 61.6 pA (19.2 to 113.9 pA), n=13 Caudal: 313.5 pA (143.8 to 729.1 pA); Mann Whitney, U=29, p=0.0035; Figure 4D]. After post-hoc staining for the cholinergic PPN and mapping the location of each recorded glutamatergic PPN neuron as described above, we identified a ‘hotspot’ of strong SNr inhibition in a medial-caudal group of glutamatergic neurons (Figure 4E). This finding shows that a subset of caudal PPN neurons preferentially receive exceptionally strong SNr input.

To determine whether the SNr input to the rostral and caudal PPN show differential short-term synaptic plasticity, we compared the normalized oIPSC amplitudes. We found that inhibitory inputs to the rostral glutamatergic neurons display short-term synaptic facilitation with amplitudes increasing in subsequent stimulations (Figure 4F). However, strong facilitation occurred in only a few neurons that drove the mean upward (Figure 4G) and there was no significant difference between the PPR observed in rostral and caudal glutamatergic neurons when activating SNr axons [median (IQR); PPR n=13 Rostral: 1.07 (0.86 to 1.38), n=13 Caudal: 0.92 (0.89 to 1.02); Mann Whitney, U=77, p=0.7241; Figure 4G].

Because glutamatergic PPN neurons spontaneously fire16,49, we wanted to determine if larger inhibitory currents mediated larger decreases in the firing rate of caudal PPN neurons. Therefore, we recorded tonic action potential firing in glutamatergic PPN neurons while optically stimulating SNr axons (Figure 4H). Although inhibition measured as the percent of pre-optical stimulation firing frequency was not significantly different between rostral and caudal neurons (mean ± SEM; %Frequency During Stimulation n=13 Rostral: 32.62 ± 10.21 %, n=17 Caudal: 24.71 ± 7.30 %; 2way ANOVA, p=0.7320; Figure 4I), we found that the absolute frequency decrease was significantly larger in caudal glutamatergic neurons compared to rostral neurons during SNr stimulation [median (IQR); ΔFrq During Opto n=13 Rostral: –3.21 Hz (–12.21 to – 1.95 Hz), n=17 Caudal: –14.08 Hz (–20.04 to –6.07 Hz); Mann Whitney, U=55, p=0.0197; Figure 4J]. To examine the discrepancy between the percent decrease and the absolute decrease in firing frequency, we compared the spontaneous firing frequency between rostral and caudal glutamatergic neurons in cells that had no holding current applied. We found that caudal glutamatergic neurons fire faster than rostral glutamatergic neurons [median (IQR); Spontaneous Frequency n=11 Rostral: 6.8 Hz (5.4 to 17.2 Hz), n=16 Caudal: 20.6 Hz (11.2 to 37.2 Hz); Mann Whitney, U=45, p=0.0343; Figure 4K]. Together, these findings show that fast-firing caudal glutamatergic PPN neurons receive larger SNr-mediated IPSCs and display greater decreases in their action potential firing.

Since we found that the caudal glutamatergic neurons receive larger inhibitory currents and display greater decreases in firing rate during SNr axon stimulation, we performed a correlation analysis to evaluate the strength of the relationship among the synaptic and firing characteristics (Figure 4L). Interestingly, there was a weak but significant positive correlation between the PPR and frequency change during stimulation showing that input displaying short-term synaptic depression occurs in neurons with greater decreases in firing frequency (Spearman r=0.486, p=0.030; Figure 4M). As expected, SNr-mediated oIPSCs with larger amplitudes are correlated with greater decreases in firing frequency during stimulation (Spearman r=-0.841, p<0.00001; Figure 4N). Greater pre-optical stimulation firing frequency is correlated with larger frequency decreases during stimulation (Spearman r=-0.791, p<0.0001; Figure 4O). This finding is expected because the fast-firing neurons can be more strongly inhibited in terms of absolute decrease in action potential frequency (i.e. a floor effect). Surprisingly, however, pre-optical stimulation firing frequency is correlated with the first oIPSC amplitude (Spearman r=0.818, p<0.0001; Figure 4P), indicating that the SNr is more strongly connected to glutamatergic PPN neurons with faster firing rates. These correlations highlight strong selectivity in SNr projections to PPN glutamatergic neurons. Overall, these findings support that the SNr selectively targets and strongly inhibits fast-firing caudal glutamatergic PPN neurons compared to rostral neurons, and that there is a medial-caudal hotspot where SNr inhibition of PPN glutamatergic neurons is particularly strong.

The SNr most strongly inhibits caudal glutamatergic PPN neurons

Most of the previous PPN circuit work has been done with rabies tracing. Aside from technical limitations, rabies tracing can only indicate the presence of synaptic connections. Our electrophysiological findings comprehensively characterize the strength and characteristics of the SNr input to each of the PPN cell types in both the rostral and caudal regions. Using this comprehensive dataset, we compared SNr input across the different PPN cell types in each region under the same electrophysiological conditions. We found that the median amplitude of the first SNr-mediated oIPSC was the same among rostral PPN neurons irrespective of cell type [median (IQR); first oIPSC n=15 ChAT+: 74.7 pA (44.5 to 136.5 pA), n=9 Vgat+: 76.5 pA (27.0 to 209.9 pA), n=13 Vglut2+: 61.6 pA (19.2 to 113.9 pA); Kruskal-Wallis, p=0.7046; Figure 5A]. The impact of these inhibitory inputs on neuronal firing did not differ with cell type as shown by the absolute change in frequency during SNr stimulation [median (IQR); ΔFrq During Opto n=14 ChAT+: –2.47 Hz (–3.37 to –1.44 Hz), n=7 Vgat+: –5.03 Hz (–5.54 to –1.34 Hz), n=13 Vglut2+: –3.21 Hz (–12.21 to –1.95 Hz); Kruskal-Wallis, p=0.3028; Figure 5B].

The SNr most strongly inhibits caudal glutamatergic PPN neurons.

(A,C) Individual cell data for the first oIPSC amplitude recorded in each cell type for rostral and caudal PPN neurons, respectively. (B,D) Individual cell data for the absolute change in frequency during stimulation in each cell type for rostral and caudal PPN neurons, respectively. (E) Graphical depiction of SNr stimulation results. * p<0.05, ** p<0.01, **** p<0.0001; box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles.

We found that the SNr-mediated oIPSCs in caudal glutamatergic neurons were significantly larger than in cholinergic and GABAergic caudal PPN neurons [median (IQR); first oIPSC n=20 ChAT+: 109.3 pA (58.5 to 168.6 pA), n=12 Vgat+: 59.7 pA (37.7 to 269.7 pA), n=13 Vglut2+: 313.5 pA (143.8 to 729.1 pA); Kruskal-Wallis, p=0.0128; Dunn’s test, ChAT+ vs Vgat+ p=0.6409, ChAT+ vs Vglut2+ p=0.0042, Vgat+ vs Vglut2+ p=0.0295; Figure 5C]. Likewise, the impact of SNr inhibition on neuron firing rate was strongest in caudal glutamatergic neurons. The absolute frequency decrease during stimulation in caudal glutamatergic neurons was larger than both cholinergic and GABAergic neurons [median (IQR); ΔFrq During Opto n=23 ChAT+: –2.10 Hz (–3.11 to – 0.84 Hz), n=19 Vgat+: –4.29 Hz (–11.02 to –2.94 Hz), n=17 Vglut2+: –14.08 Hz (–20.04 to –6.07 Hz); Kruskal-Wallis, p<0.0001; Dunn’s test, ChAT+ vs Vgat+ p=0.0049, ChAT+ vs Vglut2+ p<0.0001, Vgat+ vs Vglut2+ p=0.0332; Figure 5D].

We found no significant difference between SNr-mediated oIPSCs recorded GABAergic and cholinergic neurons in both the rostral and caudal PPN (Figure 5A&C). In the caudal PPN, however, the GABAergic neuron firing frequency was more strongly inhibited than cholinergic (Figure 5D). The greater reduction in firing frequency may be driven by the caudal GABAergic subgroup receiving particularly large oIPSCs (Figure 3D) and could be explained if GABAergic PPN neurons have higher input resistance than cholinergic neurons, as has been previously suggested57. Altogether, our findings show that the SNr most strongly inhibits the glutamatergic neurons followed by GABAergic neurons in the caudal PPN and more weakly inhibits rostral PPN neurons without any cell type bias (Figure 5E).

The GPe preferentially, but weakly, inhibits a subset of caudal GABAergic and glutamatergic PPN neurons

GPe projections to the cholinergic and glutamatergic PPN neurons have been shown in previous rabies tracing studies14–17, but projections to the GABAergic MLR neurons was not detected14. To evaluate the synaptic strength of these connections and determine regional connectivity, we repeated the previous whole-cell patch clamp experiments paired with optogenetics, but this time injecting ChR2 into the GPe of ChAT-Cre/Ai9-tdTomato, Vgat-Cre/Ai9-tdTomato, and Vglut2-Cre/tdTomato mice (Figure 6A). If oIPSCs were observed in response to optical stimulation, the neuron was identified as connected. In a subset of neurons, a GABA-a receptor blocker, GABAzine, was applied to confirm the oIPSCs were GABA-mediated (See Methods for details; Figure 6B). We recorded GPe-mediated oIPSCs in all three cell types of the PPN. As predicted by the low density of GPe axons present in the rostral PPN (Figure 1D), very few rostral PPN neurons received synaptic input from the GPe. This was true within each molecularly-defined PPN cell type [caudal vs rostral % connected; ChAT+: 30% vs 0% (n=9/30 vs 0/7), Vgat+: 68% vs 38% (n=21/31 vs 6/16), Vglut2+: 69% vs 26% (n=24/35 vs 5/19); Figure 6Ci-iii]. These findings suggest that the GPe preferentially targets caudal PPN neurons compared to rostral neurons.

GPe inhibition of the three PPN cell types.

(A) Experimental set up to identify red ChAT+, Vgat+, and Vglut2+ PPN neurons for whole-cell patch clamp while stimulating ChR2-filled GPe axons [N=6]. (B) Example trace of the first five oIPSCs in the 2-second 20 Hz train [blue] inhibited by GABA-a receptor blocker, GABAzine [green], while holding the cell at –50mV. (C) Left, Percent connected among patched neurons in the rostral and caudal regions and, right, cell mapping of patched locations with the first oIPSC amplitude represented by the color scale. Top to bottom, i. ChAT+, ii. Vgat+, and iii. Vglut2+ datasets. (D) Average oIPSC amplitude at each of 40 optogenetic light pulses in n=6 ChAT+, n=19 Vgat+, and n=15 Vglut2+ caudal PPN neurons. (E) Left, Individual cell data for the first oIPSC amplitude and, right, example current traces. (F) Normalized current amplitudes in C. (G) Left, Individual cell data for the PPR between the first two oIPSC amplitudes in the train; p=0.0206. Right, top, example current trace of short-term synaptic facilitation in VgAT+ neurons. Right, bottom, example current traces of short-term synaptic depression in Vgat+ and Vglut2+ neurons. (H) Percent of pre-optical stimulation firing frequency [%Pre-Opto Frq] during and post-stimulation in n=25 ChAT+, n=18 Vgat+, and n=29 Vglut2+ caudal PPN neurons. (I) Individual cell data for the absolute change in frequency during stimulation [ΔFrq During Opto]. (J) Correlation analysis for Vgat+ neurons, color scale representing Spearman r [-1,1] and size representing p-value [1,0]. (K) Negative correlation between the absolute change in frequency during stimulation and first oIPSC amplitude; r=-0.627, p=0.044. (L) Correlation analysis for Vglut2+ neurons. (M) Negative correlation between the absolute change in frequency during stimulation and the pre-stimulation firing frequency; r=-0.648, p=0.014. *p<0.05, **p<0.01; box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles.

We then tested if the percentage of connected caudal neurons was greater than the percentage of connected rostral neurons. While the caudal connectivity was significantly greater than rostral connectivity in GABAergic and glutamatergic neurons (one-sided Fisher’s exact test; Vgat: OR=0.2857, CI=0.09190 to 0.9391, p=0.047; Vglut2: OR=0.1637, CI=0.05069 to 0.5767, p=0.0033; Figure 6Cii&iii), we found no statistical difference between the rostral and caudal connectivity in cholinergic neurons (OR=0, CI=0 to 1.301, p=0.1150; Figure 6Ci). Because the number of connected caudal cholinergic neurons was not significantly greater than its 0% rostral connectivity (Figure 6Ci), GPe inhibition of PPN cholinergic neurons is likely weak to non-existent. Since so few rostral PPN neurons received GPe inhibition, we focused our characterization of GPe inhibition to the caudal PPN.

We then evaluated the strength and impact of GPe inhibition of the few connected caudal cholinergic PPN neurons. First, we measured the amplitude of each oIPSC in the 20 Hz train over 2 seconds. We found that the average oIPSC amplitude across the train was small (approximately 20 pA, Figure 6D). Similarly, the first oIPSC measured in the connected cholinergic neurons was also quite small [median (IQR); first oIPSC n=6 ChAT+: 22.6 pA (9.75 to 50.20 pA); Figure 6E]. Among recorded caudal cholinergic PPN neurons, both connected and non-connected, these small inhibitory currents displayed negligible effects on the neuronal firing frequency of the cholinergic PPN population in both the percent of pre-optical stimulation frequency and absolute frequency change [median (IQR); n=29 ChAT+: %Frequency During Stimulation 96.67 % (85.1 to 100.7 %); ΔFrq During Opto –0.16 Hz (–0.90 to 0.03 Hz); Figure 6H&I]. These findings indicate that the GPe has essentially no direct effect on the cholinergic PPN neural population as a whole. For this reason, we exclude GPe inhibition of cholinergic PPN neurons in the synaptic characterizations shown in Figures 6F&H. Together, we show that the GPe preferentially inhibits caudal GABAergic and glutamatergic neurons while largely avoiding rostral PPN neurons.

To determine whether the synaptic strength and characteristics from the GPe to the PPN differed between GABAergic and glutamatergic PPN subtypes, we evaluated the amplitude of the GPe-mediated oIPSCs during the 2-second 20 Hz optical stimulation. Because there were few GPe-connected neurons in the rostral PPN, all recorded neurons in these datasets were restricted to the caudal PPN for the analyses in the following paragraphs. Both GABAergic and glutamatergic neurons responded to GPe stimulation with oIPSCs ranging from ones to hundreds of picoamperes (Figure 6D). While there was no significant difference in the first oIPSC measured in GABAergic and glutamatergic neurons [median (IQR); first oIPSC n=19 Vgat+: 24.77 pA (6.72 to 54.17 pA), n=15 Vglut2+: 65.71 pA (14.50 to 133.4 pA); Kruskal-Wallis, p=0.2176; Figure 6E], the normalized amplitudes show that GPe input to GABAergic neurons facilitates with subsequent stimulations, while the input to glutamatergic neurons depresses (Figure 6F). The GPe-mediated inhibitory input to glutamatergic neurons displayed a significantly lower PPR than to GABAergic neurons [median (IQR); PPR n=19 Vgat+: 0.97 (0.76 to 1.2), n=15 Vglut2: 0.81 (0.55 to 0.93); Mann-Whitney, U=76, p=0.0206; Figure 6G]. While short-term synaptic plasticity patterns were variable among GABAergic neurons with some depressing and others facilitating, GPe input to glutamatergic neurons generally depresses.

While there was no difference in oIPSC amplitudes, short-term synaptic depression can dampen the impact that repeated IPSCs have on action potential output. Therefore, we compared the firing frequency change in each cell type during optogenetic stimulation of GPe axons. As expected, the inhibition of action potential firing in both GABAergic and glutamatergic neurons was significantly greater than the overall GPe inhibition of cholinergic neurons (mean ± SEM; %Frequency During Stimulation n=25 ChAT+: 92.39 ± 7.59 %, n=18 Vgat+: 57.70 ± 10.11 %, n=29 Vglut2+: 64.15 ± 7.63 %; 2way ANOVA, p=0.0302; Tukey test, During Stimulation ChAT+ vs Vgat+ p=0.0359, ChAT+ vs Vglut2+ p=0.0391, Vgat+ vs Vglut2+ p=0.8669; Pre vs During ChAT+ p=0.5818, Vgat+ p=0.0017, Vglut2+ p=0.0002; median (IQR); ΔFrq During Opto n=25 ChAT+: –0.16 Hz (–0.90 to 0.03 Hz), n=18 Vgat+: –1.92 Hz (–3.85 to –0.60 Hz), n=29 Vglut2+: –0.89 Hz (–3.28 to –0.43 Hz); Kruskall-Wallis, p=0.0002; Dunn’s test, ChAT+ vs Vgat+ p=0.0002, ChAT+ vs Vglut2+ p=0.0009, Vgat vs Vglut2+ p=0.4020; Figure 6H&I). The PPN GABAergic neuron population firing rate was decreased to about 58% of their pre-stimulation firing with a median frequency decrease of 1.92 Hz during stimulation (Figure 6H&I). Similarly, the glutamatergic neuron population firing rate was decreased to about 64% of their pre-stimulation firing with a median frequency decrease of 0.89 Hz during stimulation (Figure 6H&I). We found no statistical difference between GPe inhibition of firing frequency in caudal GABAergic and glutamatergic PPN neurons. These findings show that the GPe inhibits GABAergic and glutamatergic neural populations to a similar extent.

To determine if there are significant relationships among the synaptic and firing characteristics, we performed a correlation analysis. Among GABAergic neurons receiving GPe inhibition, we found that the amplitude of the first oIPSC is significantly correlated with the absolute change in frequency during stimulation (Spearman r=-0.627, p=0.044, Figure 6K). Although GPe input to GABAergic neurons exhibited various PPRs, the correlation analyses revealed no significant relationship between PPR and absolute change in frequency during stimulation (Figure 6J). These findings suggest that the initial oIPSC amplitude of GPe input to GABAergic PPN neurons has a stronger relationship with firing rate than short-term synaptic plasticity does.

As nearly all GPe inputs to glutamatergic neurons exhibited short-term synaptic depression, we expected the correlation analysis to reveal a significant relationship between PPR and the change in firing frequency. However, similar to the GPe input to GABAergic neurons, PPR was not predictive of impact on neuronal activity (Figure 6L). GPe inhibition of the glutamatergic neurons showed a significant correlation between the absolute change in frequency during stimulation and pre-stimulation firing frequency (Spearman r=-0.648, p=0.014, Figure 6M). However, because the inhibitory synaptic current amplitude did not correlate with firing frequency, this correlation is likely due to a floor effect in which faster firing neurons have a greater capacity for reductions in firing frequency. Together, these findings suggest that the GPe preferentially inhibits caudal GABAergic and glutamatergic PPN neurons.

In vivo activation of GPe and SNr axons in the PPN show opposite effects on locomotion and valence

Once we established that the SNr and GPe inhibit different profiles of regionally– and molecularly-defined subpopulations of the PPN using ex vivo optogenetic electrophysiology, we wanted to evaluate the behavioral consequences of selectively stimulating these inputs in vivo. Direct optogenetic activation of PPN subpopulations have been previously shown to either promote or inhibit motion, sometimes with contradictory results14,16,17,22–27. To determine whether differential inhibitory inputs onto the PPN influenced locomotor activity, we bilaterally injected wild-type mice with ChR2 in either the SNr or GPe and implanted an optical fiber over the PPN (Figure 7A&B). Control mice were injected with an EGFP virus and implanted. After three weeks, we placed each mouse in the open field to measure gross locomotor behavior with and without optical stimulation of the SNr or GPe axons. We tracked the distance traveled for each mouse at baseline and during bilateral stimulation of either the SNr or GPe by applying blue light (473 nm, 4-4.5 mW) at 20 Hz with 2 ms pulses (4% duty cycle), for 1 minute at a time (Figure 7C). We found that stimulation of these two inhibitory basal ganglia inputs to the PPN resulted in completely opposite motor behaviors. Specifically, stimulation of SNr axons in the PPN increased distance traveled, while stimulation of GPe axons in the PPN decreased distance traveled [mean ± SEM; N=9 (3M, 6F) Control (Ctrl): 2.22 ± 0.24 m, N=8 (5M, 3F) SNr: 4.11 ± 0.51 m, N=9 (4M, 5F) GPe: 0.57 ± 0.18 m; Welch ANOVA, Dunnett’s test, p<0.0001; Ctrl vs SNr p=0.0138, Ctrl vs GPe p=0.0001; Figure 7D, see Video S1&2].

In vivo activation of GPe and SNr axons in the PPN show opposite effects on locomotion and valence.

(A) Experimental set up to stimulate ChR2-filled SNr or GPe axons over the PPN in vivo. (B) Representative image of optical fiber tract overlaid with the approximate optical fiber placement for SNr-[green] and GPe-[orange] injected mice. (C) Distance traveled over time in an open field with 1 minute 20 Hz optical stimulations over the PPN in N=9 control (Ctrl) mice (black circles), N=8 mice injected with ChR2 in the SNr (green diamonds), and N=9 mice injected with ChR2 in the GPe (orange hexagons); R=recovery period. (D) Average distance traveled for each mouse across the six 1minute optical stimulations. (E) Representative mouse track tracings during real time place preference task in a three-chamber box and continuously stimulating EGFP-or ChR2-filled axons over the PPN at 20 Hz in SNr– and GPe-injected mice when the mice are in the stimulation zone. (F) Percent time spent in the stimulation zone in N=16 control mice, N=9 mice injected with ChR2 in the SNr, and N=10 mice injected with ChR2 in the GPe. (G) Average distance traveled in the open field during optical stimulations and (H) percent time spent in the stimulation zone using 0.25 mW laser power in N=6 mice injected with ChR2 in the SNr and N=8 mice injected with ChR2 in the GPe. * p<0.05, ** p<0.01, *** p<0.001, **** p<0.0001; box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles. See related supplemental Video S1-S4.

The PPN has been implicated in reward processing and stimulation of the PPN can be reinforcing7,27,28,30,34,58. Therefore, we evaluated the effect of GPe or SNr axon stimulation in the PPN on valence in the real-time place preference (RTPP) task. In a three-chamber apparatus, the mouse could freely move between chambers for ten minutes. Optical stimulation of the SNr or GPe axons in the PPN was applied when the mouse entered the stimulated chamber and remained on at 20 Hz (4% duty cycle) until the mouse exited that chamber (Figure 7E). To reduce the effects of optical stimulation on locomotion, this experiment was conducted with unilateral stimulation only. Interestingly, we again found opposite effects when stimulating the two inhibitory basal ganglia inputs to the PPN. Mice avoided the stimulated chamber when the SNr axons were stimulated, but preferred the stimulated chamber when the GPe axons were stimulated (mean ± SEM; N=16 (6M, 10F) Ctrl: 40.76 ± 2.0 %, N=9 (5M, 4F) SNr: 23.99 ± 5.2 %, N=10 (5M, 5F) GPe: 76.11 ± 3.7 %; One-way ANOVA, p<0.0001; Dunnett’s test, Ctrl vs SNr p=0.0025, Ctrl vs GPe p<0.0001; Figure 7F, see Video S3&4).

In addition to sending inhibitory inputs to the PPN, the GPe and SNr send axon collaterals to multiple other brain areas12,36,59,60. Therefore, it is possible that stimulating these axons in the PPN could cause antidromic stimulation of the GPe and SNr cell bodies, resulting in the inhibition of non-PPN brain regions. However, it has recently been shown that 0.25 mW laser power stimulation of axons can prevent antidromic activation of cell bodies41. Therefore, we repeated these optogenetic behavioral experiments using 0.25 mW laser power. We found that the locomotion effect of SNr stimulation over the PPN was no longer present during low power stimulation, but GPe stimulation still significantly decreased locomotion (mean ± SEM; distance traveled N=9 (3M, 6F) Ctrl: 2.22 ± 0.24 m, N=6 (3M, 3F) SNr: 2.24 ± 0.50 m, N=8 (4M, 4F) GPe: 1.09 ± 0.22 m; One-way ANOVA, p=0.0211; Dunnett’s test, Ctrl vs SNr p=0.9993, Ctrl vs GPe p=0.0232; Figure 7H). In the RTPP task, SNr axon stimulation continued to evoke place aversion and GPe axon stimulation continued to evoke place preference (mean ± SEM; % time in stimulation zone N=16 (6M, 10F) Ctrl: 40.76 ± 2.1 %, N=6 (3M, 3F) SNr: 27.72 ± 4.0 %, N=8 (4M, 4F) GPe: 68.41 ± 8.5 %; Welch’s ANOVA, p=0.0044; Dunnett’s test, Ctrl vs SNr p=0.0378, Ctrl vs GPe p=0.0252; Figure 7G). These findings show that decreased locomotion and place preference are mediated by GPe inhibition of the PPN. Interestingly, we find that aversion is mediated by SNr inhibition of the PPN, but the enhancement of locomotion is likely due to PPN-projecting SNr neurons inhibiting a separate brain structure.

Discussion

In this study, we comprehensively characterize the synaptic strength and impact on neuron action potential firing in molecularly– and regionally-defined PPN subpopulations while stimulating the SNr or GPe using ex vivo electrophysiology and optogenetics. We also stimulated the SNr or GPe axons in the PPN in vivo which revealed SNr and GPe populations, defined by their projections to the PPN, involved in non-canonical basal ganglia circuits to evoke opposing locomotion and valence processing behaviors. Together, these findings show that distinct and selective inhibition of PPN subpopulations by the SNr and GPe can alter behavioral output.

Region– and cell type-specific inputs to the PPN

The SNr and GPe have been previously shown to form monosynaptic synapses with PPN neurons using rabies viral tracing12–17. While SNr input to cholinergic and glutamatergic PPN neurons has been reproduced across different studies13–17, SNr inputs to GABAergic PPN neurons has only been inferred from studies of the entire mesencephalic locomotor region (MLR), consisting of both the PPN and cuneiform nucleus14. Several studies have used slice electrophysiology to show that the SNr inhibits PPN neurons generally31 or to compare SNr input to cholinergic and non-cholinergic PPN neurons47,50,51. However, a full characterization of the SNr input to regionally– and molecularly-defined PPN neurons in adult animals has been lacking. Using ex vivo whole-cell patch recordings, we were able to measure inhibitory currents in each of the PPN cell types during SNr stimulation showing that the SNr functionally inhibits each cell type within the rostral PPN to a similar extent while most strongly inhibiting a medial-caudal ‘hotspot’ of glutamatergic neurons.

While the GPe has been shown to project to the cholinergic and glutamatergic PPN neurons15–17, our electrophysiological results show that GPe inhibition of the cholinergic neurons is weak to non-existent and its inhibition of the glutamatergic PPN neurons is caudally biased. In contrast to a previous study showing that the GPe does not project to GABAergic MLR neurons14, we find that the GPe also preferentially inhibits a caudal subgroup of GABAergic PPN neurons. While the GPe selectively inhibits a caudal subpopulation of GABAergic and glutamatergic PPN neurons, its inhibition is much weaker than SNr inhibition of the PPN in terms of the proportion of connected PPN neurons, the inhibitory current amplitudes, and the inhibition of neuronal firing. Overall, our findings utilize a systematic approach to characterize the synaptic strength between these inhibitory basal ganglia nuclei and the PPN across its rostrocaudal axis for each cell type allowing us to compare SNr and GPe to PPN circuitry with regional-level granularity. These results also encourage future work involving the PPN to consider the individual influence of its rostral and caudal regions.

Noncanonical basal ganglia motor circuits

We found that stimulation of GPe axons over the PPN decreased locomotion in both low (local) and high (putatively generating antidromic activity) laser power stimulations, while only high power SNr axon stimulation increased locomotion. This suggests that antidromically activating the subpopulation of SNr neurons that project to the PPN is required to generate the non-canonical increase in locomotion, but that this locomotion increase is not due to the SNr inhibition of the PPN. These behavioral outcomes appear counter to the canonical model of basal ganglia movement pathways, in which SNr activation decreases and GPe activation increases movement. Although there is strong support for this canonical model, recent in vivo recordings have found subpopulations of SNr and GPe neurons that show activity patterns counter to this model (i.e., SNr neurons that increase activity during movement and GPe neurons that increase activity during immobility)61–67. Similarly, a subset of SNr and GPe neurons paradoxically increase activity upon direct and indirect pathway striatal projection neuron stimulation, respectively40,68. While one explanation for these heterogeneous responses could be local inhibition within each structure39,40,69–73, another possibility is that distinct subpopulations within the SNr and GPe differentially modulate locomotor behavior12,33,38–40,59,63,71. Our findings support non-canonical basal ganglia motor pathways involving SNr and GPe neurons that project to the PPN.

Our characterization of region-specific inhibitory inputs to the PPN helps us understand how GPe inhibition of the PPN can mediate decreases in locomotion. Our electrophysiology experiments show that the GPe exerts minimal influence on the rostral PPN neurons which appear to decrease locomotion22,24 and preferentially inhibits caudal GABAergic and glutamatergic PPN neurons. Because stimulating the caudal GABAergic and glutamatergic PPN neurons increases locomotion23, GPe selective inhibition of these caudal neurons could explain our behavioral experiments showing that stimulation of GPe axons over the PPN decreases locomotion. Aligned with our findings, selectively stimulating the Npas1+ or FoxP2+ subpopulations in the GPe decreases locomotion38,40. Some groups have found that direct stimulation of the PV+ GPe neurons increases locomotion38,60; however, one study shows that inhibition of PV+ GPe neurons can increase locomotion41. While we show that GPe projections to the PPN decrease locomotion, future studies are needed to determine which GPe subpopulations project to the PPN to modulate locomotion.

While we found that stimulating SNr axons in the PPN with high laser power increased locomotion, we observed no locomotor effect when using low laser power stimulation. Therefore, our findings suggest that antidromic activation of the PPN-projecting SNr neuron subpopulation during high laser power stimulation promotes locomotion, but not through direct actions at the PPN. Previous studies have identified distinct functional roles for molecularly-defined subpopulations within the SNr. Although direct stimulation of PV+ and GAD2+ SNr neurons decreases gross locomotor movement33,63, PV+ SNr neurons have been shown to increase their activity during transitions from quiet wakefulness to non-locomotor movement63. The subset of PV+ and GAD2+ SNr neurons which project to the PPN also broadly project to other SNr output targets12,63 whose inhibition may mediate increases in locomotion. However, stimulation of SNr neurons has not previously been shown to increase locomotion. One possibility is that the adjacent dopaminergic neurons of the substantia nigra pars compacta (SNc) are infected with ChR2 in our experiments. Some SNc dopaminergic neurons send axons to both the PPN and striatum74. Because direct stimulation of nigrostriatal neurons increases locomotion75, antidromic stimulation of the SNc can evoke dopamine release into the striatum to increase movement in our high laser power experiments. However, we also find that this high-power SNr axon stimulation is highly aversive (Figure 7F). Because dopaminergic neuron activation is usually rewarding, our aversive effect does not support antidromic activation of these dopaminergic neurons. Our findings suggest the existence of a SNr subpopulation, defined by their projections to the PPN, that can positively modulate movement. However, future work is needed to test the locomotor effects of direct stimulation of PPN-projecting SNr neurons.

Differential inhibition of PPN subpopulations by the SNr and GPe mediate opposing valence processing outcomes

We found that stimulation of SNr axons over the PPN was aversive in real time place preference (RTPP), while stimulation of GPe axons over the PPN led to increased preference for the stimulated chamber. Previous studies directly stimulating SNr and GPe subpopulations have implicated these structures in reward and aversion processing. Inhibition of Vgat+ SNr neurons has been shown to increase place preference35 while its direct stimulation has no effect33,35. However, selective stimulation of the PV+ SNr neurons induces aversion in RTPP33. Although the GPe has been predominantly studied in the context of movement, recent work has begun to reveal a role for the GPe in valence processing41,76–78. One recent study shows that inhibition of PV+ GPe neurons induced aversion while inhibition of Npas1+ GPe neurons induced placed preference in RTPP41. Because activation of PV+ SNr neurons is aversive33 and inhibition of PV+ GPe neurons is aversive41, it is likely that our results are due to activation of the PV+ SNr and GPe axons in the PPN. While we show that activation of the PPN-projecting SNr and GPe subpopulations strongly affect valence processing, future studies are needed to determine whether these effects are specifically mediated by PV+ SNr and GPe projections to the PPN.

Neurons in the PPN have been implicated in reward and aversive-related behaviors7,27,28,30–32,34,58. The cholinergic PPN neurons are implicated in positive valence with place preference and increased lever pressing through their projections to the ventral tegmental area (VTA)27,28. Aligned with our findings, direct inhibition of the cholinergic PPN neurons evokes place aversion27. Of particular significance, we show that the SNr inhibits the cholinergic PPN neurons while the GPe axons avoid cholinergic PPN neurons. Therefore, it is possible that the difference between the SNr and GPe effects on valence is due to their differential inhibition of the cholinergic PPN neurons.

Direct stimulation of the glutamatergic PPN neurons has also been implicated in positive valence with increased reinforcement behavior7. Both the cholinergic and glutamatergic PPN neurons directly influence dopamine release in the striatum and dopaminergic neuron activity in the SNc and VTA7,27,28,79–83. Specifically, the caudal PPN neurons are thought to innervate the medial part of the SNc and the VTA9,83, two major reward pathway hubs. Our data show that the SNr most strongly inhibits the caudal glutamatergic PPN neurons. Therefore, strong SNr inhibition of these neurons would remove excitatory drive from the reward-related dopaminergic neurons. This pathway could also contribute to the aversive effect of SNr axon stimulation in the PPN.

The pathways underlying the GPe-mediated rewarding effect are less clear and may indicate functional heterogeneity among caudal PPN neurons. We find that the GPe only inhibits a subset of caudal GABAergic and glutamatergic PPN neurons. It is possible that the subset of GABAergic PPN neurons targeted by the GPe disinhibits local cholinergic and glutamatergic neurons to increase excitation of the rewarding dopaminergic neuron pathways. Another possibility is that the specific GPe-inhibited PPN population may have particularly aversive-properties. For example, PPN neurons that excite amygdala nuclei evoke place avoidance34,58. In this case, GPe inhibition of these neurons could be rewarding by removing excitation of amygdala nuclei involved in negative valence processing. Future experimental work is needed to determine the extent of local inhibition among PPN neurons and define the characteristics of the PPN neurons selectively targeted by the GPe.

Conclusions

We have systematically characterized SNr and GPe inputs across the rostrocaudal axis of the PPN for each cell type – cholinergic, GABAergic, and glutamatergic. We show that the SNr inhibits nearly every PPN cell recorded with differential strength determined by cell type and anatomical biases while identifying a medial-caudal ‘hotspot’ of glutamatergic neurons most strongly inhibited by the SNr. In contrast, the GPe strikingly avoids the cholinergic PPN neurons and more weakly, but selectively, inhibits a subpopulation of caudal GABAergic and glutamatergic neurons. We proposed that the differential inhibition of regionally-biased, mixed-cell type PPN subpopulations can alter behavioral outputs. By stimulating these inhibitory basal ganglia axons over the PPN, we show that the SNr evokes place aversion while the GPe evokes place preference and decreases locomotion. Surprisingly, we find that high laser power stimulation (likely to evoke antidromic activation) of PPN-projecting SNr neurons increases locomotion. Together, our findings show that the SNr and GPe mediate opposing valence processing outcomes through the PPN and support non-canonical basal ganglia motor pathways in which the PPN-projecting SNr subpopulation increases locomotion and the GPe decreases locomotion through its projections to the PPN.

Supporting information

eLife Assessment

Fallah et al carefully dissect projections from substantia nigra pars reticulata (SNr) and the globus pallidus externa (GPe) – two key basal ganglia nuclei – to the pedunculopontine nucleus (PPN), a brainstem nucleus that has a central role in motor control. They consider inputs from these two areas onto 3 types of downstream PPN neurons – GABAergic, glutamatergic, and cholinergic neurons – and carefully map connectivity along the rostrocaudal axis of the PPN. Overall, this valuable study provided convincing data on PPN connectivity with two key input structures that will provide a basis for further understanding PPN function.

Reviewer #1 (Public review):

Summary:

Fallah and colleagues characterize the connectivity between two basal ganglia output nuclei, the SNr and GPe, and the pedunculopontine nucleus, a brainstem nucleus that is part of the mesencephalic locomotor region. Through a series of systematic electrophysiological studies, they find that these regions target and inhibit different populations of neurons, with anatomical organization. Overall, SNr projects to PPN and inhibits all major cell types, while the GPe inhibits glutamatergic and GABAergic PPN neurons, and preferentially in the caudal part of the nucleus. Optogenetic manipulation of these inputs had opposing effects on behavior - SNr terminals in the PPN drove place aversion, while GPe terminals drove place preference.

Strengths:

This work is a thorough and systematic characterization of a set of relatively understudied circuits. They build on the classic notions of basal ganglia connectivity and suggest a number of interesting future directions to dissect motor control and valence processing in brainstem systems.

Weaknesses:

Characterization of the behavioral effects of manipulations of these PPN input circuits could be further parsed, for a better understanding of the functional consequences of the connections demonstrated in the ephys analyses.

All the cell type recording studies showing subtle differences in the degree of inhibition and anatomical organization of that inhibition suggest a complex effect of general optogenetic manipulation of SNr or GPe terminals in the PPN. It will be important to determine if SNr or GPe inputs onto a particular cell type in PPN are more or less critical for how the locomotion and valence effects are demonstrated here.

Reviewer #2 (Public review):

Summary:

Fallah et al carefully dissect projections from SNr and GPe - two key basal ganglia nuclei - to the PPN, an important brainstem nucleus for motor control. They consider inputs from these two areas onto 3 types of downstream PPN neurons: GABAergic, glutamatergic, and cholinergic neurons. They also carefully map connectivity along the rostrocaudal axis of the PPN.

Strengths:

The slice electrophysiology work is technically well done and provides useful information for further studies of PPN. The optogenetics and behavioral studies are thought-provoking, showing that SNr and GPe projections to PPN play distinct roles in behavior.

Weaknesses:

Although the optogenetics and behavioral studies are intriguing, they are somewhat difficult to fit together into a specific model of circuit function. Perhaps the authors can work to solidify the connection between these two arms of the work. Otherwise, there are a few questions whose answers could add context to the interpretation of these results:

(1) Male and female mice are used, but the authors do not discuss any analysis of sex differences. If there are no sex differences, it is still useful to report data disaggregated by sex in addition to pooled data.

(2) There is some lack of clarity in the current manuscript on the ages used - 2-5 months vs "at least 7 weeks." Is 7 weeks the time of virus injection surgery, then recordings 3 weeks later (at least 10 weeks)? Please clarify if these ages apply equally to electrophysiological and behavioral studies. If the age range used for the test is large, it may be useful to analyze and report if there are age-related effects.

(3) Were any exclusion criteria applied, e.g. to account for missed injections?

(4) 28-34degC is a fairly wide range of temperatures for electrophysiological recording, which could affect kinetics.

(5) It would be good to report the number of mice used for each condition in addition to n=cells. Statistically, it would be preferable not to assume that each cell from the same mouse is an independent measurement and to use a nested ANOVA.

Reviewer #3 (Public review):

Summary:

The study by Fallah et al provides a thorough characterization of the effects of two basal ganglia output pathways on cholinergic, glutamatergic, and GABAergic neurons of the PPN. The authors first found that SNr projections spread over the entire PPN, whereas GPe projections are mostly concentrated in the caudal portion of the nucleus. Then the authors characterized the postsynaptic effects of optogenetically activating these basal ganglia inputs and identified the PPN's cell subtypes using genetically encoded fluorescent reporters. Activation of inputs from the SNr inhibited virtually all PPN neurons. Activation of inputs from the GPe predominantly inhibited glutamatergic neurons in the caudal PPN, and to a lesser extent GABAergic neurons. Finally, the authors tested the effects of activating these inputs on locomotor activity and place preference. SNr activation was found to increase locomotor activity and elicit avoidance of the optogenetic stimulation zone in a real-time place preference task. In contrast, GPe activation reduced locomotion and increased the time in the RTPP stimulation zone.

Strengths:

The evidence of functional connectivity of SNr and GPe neurons with cholinergic, glutamatergic, and GABAergic PPN neurons is solid and reveals a prominent influence of the SNr over the entire PPN output. In addition, the evidence of a GPe projection that preferentially innervates the caudal glutamatergic PPN is unexpected and highly relevant for basal ganglia function.

Opposing effects of two basal ganglia outputs on locomotion and valence through their connectivity with the PPN.

Overall, these results provide an unprecedented cell-type-specific characterization of the effects of basal ganglia inputs in the PPN and support the well-established notion of a close relationship between the PPN and the basal ganglia.

Weaknesses:

The behavioral experiments require further analysis as some motor effects could have been averaged out by analyzing long segments. Additional controls are needed to rule out a motor effect in the real-time place preference task. Importantly, the location of the stimulation is not reported even though this is critical to interpret the behavioral effects.

There are some concerns about the possible recruitment of dopamine neurons in the SNr experiments.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Fallah and colleagues characterize the connectivity between two basal ganglia output nuclei, the SNr and GPe, and the pedunculopontine nucleus, a brainstem nucleus that is part of the mesencephalic locomotor region. Through a series of systematic electrophysiological studies, they find that these regions target and inhibit different populations of neurons, with anatomical organization. Overall, SNr projects to PPN and inhibits all major cell types, while the GPe inhibits glutamatergic and GABAergic PPN neurons, and preferentially in the caudal part of the nucleus. Optogenetic manipulation of these inputs had opposing effects on behavior - SNr terminals in the PPN drove place aversion, while GPe terminals drove place preference.

Strengths:

This work is a thorough and systematic characterization of a set of relatively understudied circuits. They build on the classic notions of basal ganglia connectivity and suggest a number of interesting future directions to dissect motor control and valence processing in brainstem systems. We thank the reviewer for these positive comments.

Weaknesses:

Characterization of the behavioral effects of manipulations of these PPN input circuits could be further parsed, for a better understanding of the functional consequences of the connections demonstrated in the ephys analyses.

We have further analyzed our behavioral data to reveal more nuanced functional effects and included these analyses in Figure S2.

All the cell type recording studies showing subtle differences in the degree of inhibition and anatomical organization of that inhibition suggest a complex effect of general optogenetic manipulation of SNr or GPe terminals in the PPN. It will be important to determine if SNr or GPe inputs onto a particular cell type in PPN are more or less critical for how the locomotion and valence effects are demonstrated here.

This is a really interesting future direction and we have expanded on these points in the discussion in lines 771-772 and 782-785.

Reviewer #1 (Recommendations for the authors):

(1) Overall these are really valuable studies and help set up a number of future directions.

We thank the reviewer for their positive comments.

(2) I don't have many specific suggestions, but more examples of viral targeting and cell type targeting, including potentially some validation of the genetic identity of the cells targeted, could be useful for considering the details of the ephys experiments.

We agree that understanding which exact SNr and GPe neurons go to which exact PPN populations is an important next step and are planning to conduct future experiments investigating these important questions. Others have found that there is minimal overlap between the three cell types within the PPN discussed in this manuscript (Wang and Morales 2009; Yoo et al. 2017; Steinkellner, Yoo, and Hnasko 2019). One important line of future investigations is to look at the specific inputs onto recently identified subsets of the glutamatergic PPN neurons such as Chx10- and Rbp4-expressing neurons (Goñi-Erro et al. 2023; Ferreira-Pinto et al. 2021). We hope to explore the electrophysiological properties and connectivity of these subtypes in future projects.

(3) More discussion of which PPN cell types might be mediating the optogenetic behavioral effects of bulk SNr or GPe terminal stimulation would be useful for connecting the ephys results with the behavior.

We are also interested in the question of which PPN cell type is most critical for mediating the effects observed in bulk terminal stimulation. While the best experiment would be to stimulate the axons projecting to each specific cell type of the PPN, this is not currently possible due to methodological limitations and lack of studies dissecting which SNr and GPe subpopulations project to each cell type of the PPN. However, in future studies, we plan to leverage the ability of AAV1 to jump a synapse along with Cre/Flp viruses and mouse lines to selectively inhibit cholinergic, GABAergic, or glutamatergic PPN neurons that receive GPe or SNr input to elucidate the contribution of each cell type in mediating behavioral changes in movement and valence processing.

To address these important future directions, we have added additional text in the discussion in lines 771-772 and 782-785.

Reviewer #2 (Public review):

Summary:

Fallah et al carefully dissect projections from SNr and GPe - two key basal ganglia nuclei - to the PPN, an important brainstem nucleus for motor control. They consider inputs from these two areas onto 3 types of downstream PPN neurons: GABAergic, glutamatergic, and cholinergic neurons. They also carefully map connectivity along the rostrocaudal axis of the PPN.

Strengths:

The slice electrophysiology work is technically well done and provides useful information for further studies of PPN. The optogenetics and behavioral studies are thought-provoking, showing that SNr and GPe projections to PPN play distinct roles in behavior.

We appreciate the reviewer’s positive evaluation.

Weaknesses:

Although the optogenetics and behavioral studies are intriguing, they are somewhat difficult to fit together into a specific model of circuit function. Perhaps the authors can work to solidify the connection between these two arms of the work.

We have expanded on these topics in the discussion.

Otherwise, there are a few questions whose answers could add context to the interpretation of these results:

(1) Male and female mice are used, but the authors do not discuss any analysis of sex differences. If there are no sex differences, it is still useful to report data disaggregated by sex in addition to pooled data.

We have added a supplementary figure (Figure S2) showing distance traveled during optical stimulation for male and female mice.

(2) There is some lack of clarity in the current manuscript on the ages used - 2-5 months vs "at least 7 weeks." Is 7 weeks the time of virus injection surgery, then recordings 3 weeks later (at least 10 weeks)? Please clarify if these ages apply equally to electrophysiological and behavioral studies. If the age range used for the test is large, it may be useful to analyze and report if there are age-related effects.

Thanks for pointing this out, we have clarified this in the methods. 7 weeks is the youngest age at which mice used for electrophysiology were injected, and all were used for electrophysiology between 2-5 months. For behavior, the youngest mice used were 11 weeks old at time of behavior (8 weeks old at injection). Mice in the GPe-stimulated condition were 110 ± 7.4 SEM days old and mice in the SNr-stimulated condition 132 ± 23.4 SEM days old. We have added these details to the revised manuscript in lines 913 and 963-964.

In addition, we have correlated distance traveled at baseline and during stimulation with age for both SN and GPe stimulated conditions. Baseline distance traveled did not correlate with age, but there was a trend toward more movement during stimulation with older mice in the SN axon stimulation group. We have included these plots in supplemental Figure S2.

(3) Were any exclusion criteria applied, e.g. to account for missed injections?

All injection sites and implant sites were within our range of acceptability, so we did not exclude any mice for missed injections or incorrect implant location.

(4) 28-34 degC is a fairly wide range of temperatures for electrophysiological recording, which could affect kinetics.

This is an important consideration, and we agree the wide temperature is not optimal. We have plotted our main measurement of current amplitude in the condition where we found significant differences between rostral and caudal PPN (SNr to Vglut2 PPN neurons) against temperature and found no correlation (Pearson’s r value = -0.0076). Similarly, we found no correlation between baseline (pre-opto) firing frequency and temperature (r = -0.068). See Author response image 1.

(5) It would be good to report the number of mice used for each condition in addition to n=cells. Statistically, it would be preferable not to assume that each cell from the same mouse is an independent measurement and to use a nested ANOVA.

For electrophysiology, the number of mice used in each experiment was 6 (3 male, 3 female). In the manuscript ‘N’ represents number of mice and ‘n’ represents number of cells. Because of the unpredictability of how many healthy cells can be recorded from one mouse, our data were planned to be collected with n=cells, and are underpowered for a nested ANOVA.

However, in many cases, rostral and caudal data were collected from the same mice. While we do not have sufficient paired data for each electrophysiological parameter, analyzing one of our main and most important findings with a paired comparison (with biological replicates being mice) shows a statistically significant difference in the inhibitory effect of SNr axon stimulation on firing rate between rostral and caudal glutamatergic neurons (p=0.031, Wilcoxon signed rank test). See Author response image 2.

Reviewer #3 (Public review):

Summary:

The study by Fallah et al provides a thorough characterization of the effects of two basal ganglia output pathways on cholinergic, glutamatergic, and GABAergic neurons of the PPN. The authors first found that SNr projections spread over the entire PPN, whereas GPe projections are mostly concentrated in the caudal portion of the nucleus. Then the authors characterized the postsynaptic effects of optogenetically activating these basal ganglia inputs and identified the PPN's cell subtypes using genetically encoded fluorescent reporters. Activation of inputs from the SNr inhibited virtually all PPN neurons. Activation of inputs from the GPe predominantly inhibited glutamatergic neurons in the caudal PPN, and to a lesser extent GABAergic neurons. Finally, the authors tested the effects of activating these inputs on locomotor activity and place preference. SNr activation was found to increase locomotor activity and elicit avoidance of the optogenetic stimulation zone in a real-time place preference task. In contrast, GPe activation reduced locomotion and increased the time in the RTPP stimulation zone.

Strengths:

The evidence of functional connectivity of SNr and GPe neurons with cholinergic, glutamatergic, and GABAergic PPN neurons is solid and reveals a prominent influence of the SNr over the entire PPN output. In addition, the evidence of a GPe projection that preferentially innervates the caudal glutamatergic PPN is unexpected and highly relevant for basal ganglia function.

Opposing effects of two basal ganglia outputs on locomotion and valence through their connectivity with the PPN.

Overall, these results provide an unprecedented cell-type-specific characterization of the effects of basal ganglia inputs in the PPN and support the well-established notion of a close relationship between the PPN and the basal ganglia.

We thank the reviewer for their positive comments.

Weaknesses:

The behavioral experiments require further analysis as some motor effects could have been averaged out by analyzing long segments.

We have further analyzed our motor effects and included these analyses in supplemental figure S2 in the revised manuscript.

Additional controls are needed to rule out a motor effect in the real-time place preference task.

To address this comment, we analyzed the second day of RTPP, where no stimulation was applied in either chamber. Specifically, we evaluated the time spent in the stimulated chamber during the first minute of the unstimulated RTPP task. We found that the mice that had SNr axon stimulation still avoided the previously stimulated chamber and the mice that had GPe axon stimulation still preferred the previously stimulated chamber. These data have been added to Figure 7 and in the results section lines 564-575.

Importantly, the location of the stimulation is not reported even though this is critical to interpret the behavioral effects.

The implant locations were generally over the middle-to-rostral PPN and we will clarify this in the revised manuscript. These locations are shown in figure 7B.

There are some concerns about the possible recruitment of dopamine neurons in the SNr experiments.

We have added experiments stimulating the SNc dopaminergic neuron axons in the PPN and found very interesting behavioral effects. These are described in more detail below and in the results lines 595-624. These data are also included in Figure S3.

Reviewer #3 (Recommendations for the authors):

(1) Locomotor activity should be analyzed as trial averages instead of session averages. The effect of SNr on locomotion might be showing a rebound of activity in cholinergic neurons, which innervate dopamine neurons and induce locomotion. Furthermore, the variability between animals should be reported, Figure 7C doesn't show a standard deviation.

This is an important point and could reveal different early and late effects of basal ganglia axon stimulation. We have added a time course graph of the trial averages for the distance traveled in the open field with higher temporal resolution (10s vs 1min). This is included in supplemental Figure S2A&B.

The variability between animals was shown as shaded area, but was too light and transparent so it was difficult to see in Figure 7C. We have changed this shading to error bars for better visibility.

(2) SNc projects to the PPN. It has recently been shown that PPN neurons respond robustly to dopaminergic activation, including effects on motor activity (Juarez Tello et al., 2024). The transductions shown in Figure S1 clearly cover to entirety of the SNc. Dopamine blockers should be used in the ex vivo experiments to rule out dopaminergic effects.

This is an important point and one we were particularly interested in as far as the behavioral experiments. We thank the reviewer for bringing this up because it led us to a really interesting result. We have now run an additional experiment using DAT-cre mice and a cre-dependent ChR2 using the same injection site at our constitutive ChR2 experiments. We found that selectively stimulating the SNc dopaminergic axons replicates the increased locomotion at high laser power and replicates the no change in locomotion at low laser power as seen with our constitutive ChR2 experiment. However, the selective dopaminergic axon activation in the PPN is rewarding at both high and low power, while the constitutive ChR2 activation is aversive. We have added these data to supplemental figure S3, and have added text in the results (lines 595624) and discussion (lines 695-734) about this new exciting finding.

While we can’t exclude the possibility of dopamine influence on the electrophysiology experiments (via changes in input resistance or channel properties), the fast synaptic currents measured are uncharacteristic of inhibitory D2 receptor currents (which would be slow), and are inhibited by the GABAa receptor blocker, GABAzine.

(3) Activation of glutamatergic neurons in the caudal PPN elicits locomotion while the same stimulation in the rostral PPN terminates locomotion. In line with this, the authors report important differences in glutamatergic neurons in the rostral vs caudal PPN (Fig. 5). For the behavioral experiments, the location of the optic fiber is not reported. This is essential for the interpretation of the behavioral experiments. Based on the recent literature, inhibiting glutamatergic neurons in the rostral and in the caudal PPN will produce opposing effects.

We absolutely agree the rostral and caudal PPN differences are functionally important. In Figure 7B, we have mapped the location of the optical fiber tip for each experiment. Our implant location was generally in the rostral-middle part of the PPN and we have added this to the methods section of the revised manuscript in lines 887 and 1048. While we did not have many implant locations that were specifically rostral or specifically caudal, we did evaluate the behavioral response for our most rostrally-located implant and our most-caudally located implant in the SN axon stimulation experiment. We found that low-power laser activation of nigral axons in the most rostral implant resulted in increased locomotion but in the most caudal implant resulted in decreased locomotion. This increased locomotion exactly what we would expect when rostral PPN neurons (that normally inhibit movement) are preferentially inhibited, and decreased locomotion is what we would expect when caudal PPN neurons (that normally promote movement) are inhibited. Future experiments using more precise rostral and caudal implant locations will be needed to fully parse out the functional role of rostral vs caudal PPN. See Figure S4 (two green implant sites are circled for one mouse because the implants were bilateral).

(4) Even though the authors made an effort to dissect out the motor component during the RTPP task, this was not entirely achieved. Low laser power was still able to decrease activity following GPe stimulation, causing the animal to spend more time in the stimulated compartment. It is not clear the reason for using RTPP as opposed to CPP, which will not have the confound of the effects on motor activity. The interpretation of these data is problematic.

This is an important consideration, and the reviewer is correct that we can’t completely eliminate a motor contribution to our RTPP experiment. We attempted to minimize potential motor confounds by utilizing unilateral stimulation and our supplemental videos show that the mice can escape the stimulated chamber.

However, to address this comment, we analyzed the second day of RTPP, where no stimulation was applied in either chamber. Specifically, we evaluated the time spent in the stimulated chamber during the first minute of the unstimulated RTPP task. We found that the mice that had SN stimulation still avoided the previously stimulated chamber and the mice that had GPe axon stimulation still preferred the previously stimulated chamber. These data have been added as Figure 7G and in the results section lines 564-575.

(5) The resting membrane potential for cholinergic, glutamatergic, and GABAergic neurons is not reported.

Since a majority of PPN neurons are spontaneously active, we have reported the average membrane voltage during the pre-optical stimulation period in supplementary table 1.

(6) During the RTPP, the animals were stimulated unilaterally with the purpose of reducing the optogenetic effects on locomotion, but no data support this claim. Please report the locomotor measurements during unilateral stimulation.

To address this comment, we have analyzed the speed of the mouse in each compartment (stimulated vs non-stimulated) during the RTPP task. We found that the mean speed does differ, in the direction expected (i.e., mice are on average slower in the GPe stimulated zone where they spend more time, and mice are on average faster in the SNr stimulated zone where they spend less time). This is expected because when the mouse spends more time in a zone, it is more likely to spend time grooming or staying still, but it could still be evidence of motor response to the stimulation. To evaluate how fast the mouse is able to move with and without unilateral stimulation, we measured maximum speed in the stimulated and unstimulated zone. We found that maximum speed does not differ between stimulated and unstimulated zones in either the SNr or GPe group. See Author response image 3.

(7) Given the similarity of the parameters evaluated for all three PPN cell types, the results could be presented in a table, it will be easier to summarize.

This is a good point and we have added supplemental tables 1-4 for key electrophysiological findings.

(8) The text is repetitive in some parts.

We have gone through the results to edit out repetitive text. For example, lines 244-260 and 274-287 have been rewritten for clarity and efficiency.

(9) Lines 609-620: the behavioral effects after SNr stimulation are not mediated by the PPN, please correct.

We have corrected this.

(10) The number of patched GABAergic neurons in the caudal PPN is almost double the number of patched neurons in the rostral PPN. This contrasts with the high density of GABAergic neurons in the rostral PPN reported in the literature, and therefore, the probability of recording GABAergic neurons will be much higher in the rostral PPN. Please comment.

It is true that there are more GABA neurons in the rostral region, but on a sagittal slice, the rostral region occupies a smaller area compared to caudal and there is a notable cluster of GABAergic neurons in the caudal region (Mena-Segovia et al. 2009). The number of visible and healthy cells with obvious fluorescence against background fluorescence in the heavily myelinated tissue of the PPN is unpredictable and it is possible that the dense number of GABA neurons in the rostral region conglomerates the fluorescence of individual cell somas, making it difficult to detect as many rostral neurons. While we did our best to equally patch rostral and caudal neurons based on our best judgment during the experiment, neurons were ultimately designated as ‘rostral’ or ‘caudal’ after post-hoc staining for the cholinergic neurons, as described below.

(11) Describe how the rostral and caudal PPN regions were defined and how the authors ensured consistency across recordings.

We have added more details about the definition of rostral vs caudal PPN in to the methods in lines 1042-1053.

(12) Please report the proportion of GABAergic neurons showing STD vs STP for rostral and caudal PPN. The data in Figure 3 might be averaging out some important differences. Figure 3L suggests some differences in the proportions.

The variability within the GABAergic population was really interesting and we plan to pursue this in the future. We have defined STD as PPR<0.95 and STP as PPR>1.05 and added the proportions of caudal and rostral GABAergic PPN neurons with each type of short-term synaptic plasticity to lines 253-257.

(13) Please report whether the mice’s compartment preferences during the habituation were taken into account for the selection of the laser-on compartment.

Mice were not habituated to the chamber in the unstimulated condition prior to the RTPP experiment. Laser-on side was randomly chosen and counter-balanced between mice. Mice were also randomly assigned to have low laser power RTPP first or high laser power RTPP first. In each case, mice were given an unstimulated 10-minute trial on the day between the first and second RTPP experiment to ‘unlearn’ which side was stimulated and the second RTPP experiment stimulated the opposite chamber compared to the first RTPP experiment. For example, one mouse would have high power stimulation on the striped side on day 1, no stimulation on day 2, and low power stimulation on the spotted side on day 3. This is now explained more thoroughly in lines 564-575 and lines 992-998.

(14) Some references to figure panels are missing in the text.

We have carefully reviewed the manuscript to ensure figure panels are referenced in the text.

(15) The interpretation in lines 724-725 is not supported by the data given that GPe inputs to cholinergic neurons are negligible.

We have reworded much of the discussion.

(16) Some parts of the discussion should go into the “ideas and speculation” subsection of the discussion.

We have rewritten sections of the discussion.

References:

Ferreira-Pinto, Manuel J., Harsh Kanodia, Antonio Falasconi, Markus Sigrist, Maria S. Esposito, and Silvia Arber. 2021. “Functional Diversity for Body Actions in the Mesencephalic Locomotor Region.” Cell 184 (17): 4564-4578.e18. https://doi.org/10.1016/j.cell.2021.07.002.

Goñi-Erro, Haizea, Raghavendra Selvan, Vittorio Caggiano, Roberto Leiras, and Ole Kiehn. 2023. “Pedunculopontine Chx10+ Neurons Control Global Motor Arrest in Mice.” Nature Neuroscience 26 (9): 1516–28. https://doi.org/10.1038/s41593-023-01396-3.

Mena-Segovia, J., B. R. Micklem, R. G. Nair-Roberts, M. A. Ungless, and J. P. Bolam. 2009. “GABAergic Neuron Distribution in the Pedunculopontine Nucleus Defines Functional Subterritories.” The Journal of Comparative Neurology 515 (4): 397–408. https://doi.org/10.1002/cne.22065.

Steinkellner, Thomas, Ji Hoon Yoo, and Thomas S. Hnasko. 2019. “Differential Expression of VGLUT2 in Mouse Mesopontine Cholinergic Neurons.” eNeuro, July. https://doi.org/10.1523/ENEURO.0161-19.2019.

Wang, Hui-Ling, and Marisela Morales. 2009. “Pedunculopontine and Laterodorsal Tegmental Nuclei Contain Distinct Populations of Cholinergic, Glutamatergic and GABAergic Neurons in the Rat.” The European Journal of Neuroscience 29 (2): 340–58. https://doi.org/10.1111/j.1460-9568.2008.06576.x.

Yoo, Ji Hoon, Vivien Zell, Johnathan Wu, Cindy Punta, Nivedita Ramajayam, Xinyi Shen, Lauren Faget, Varoth Lilascharoen, Byung Kook Lim, and Thomas S. Hnasko. 2017. “Activation of Pedunculopontine Glutamate Neurons Is Reinforcing.” The Journal of Neuroscience: The Official Journal of the Society for Neuroscience 37 (1): 38–46. https://doi.org/10.1523/JNEUROSCI.3082-16.2016.

Version 2

Abstract

Abstract

The canonical basal ganglia model predicts that the substantia nigra pars reticulata (SNr) and the globus pallidus externa (GPe) will have specific effects on locomotion: the SNr inhibiting locomotion and the GPe enhancing it. In this study, we use in vivo optogenetics to show that the GPe exerts non-canonical effects on locomotion while the SNr has no gross motor impact through inhibition of the PPN. We also show that these structures mediate opposing effects on reward. To understand how these structures differentially modulate the PPN, we use ex vivo whole-cell recording with optogenetics to comprehensively dissect the SNr and GPe synaptic connections to regionally- and molecularly-defined populations of PPN neurons. The SNr inhibits all PPN subtypes, but most strongly inhibits caudal glutamatergic neurons. The GPe selectively inhibits caudal glutamatergic and GABAergic neurons, avoiding both cholinergic and rostral cells. This circuit characterization reveals non-canonical basal ganglia pathways for locomotion and valence.

Introduction

The pedunculopontine nucleus (PPN) is a brainstem structure heavily interconnected with subcortical structures, such as the basal ganglia, thalamus, and spinal cord. The PPN has been implicated in locomotor control and valence processing1–4. However, understanding PPN circuitry is complicated by its anatomical and molecular heterogeneity. It has three molecularly-defined cell types, distinguished by their major neurotransmitter5 and distinct rostral and caudal topography. While cholinergic and glutamatergic neurons are more densely packed in the caudal PPN, GABAergic neurons are more densely packed in the rostral PPN5,6. These molecularly-defined populations display very minimal overlap5,7,8. This unique topography has been conserved across species9–11, but it is unknown how the basal ganglia connects to each of the distinct cell types in the rostral and caudal PPN regions to modulate locomotor and valence outputs.

While the PPN interacts with all basal ganglia nuclei3,9, the inhibitory inputs from the substantia nigra pars reticulata (SNr) and the globus pallidus externus (GPe) to the PPN are of particular interest because of their roles in the canonical motor output pathways for ‘stop’ and ‘go,’ respectively. Both the SNr and GPe are known to send axonal projections to the PPN12–18. Previous studies have identified SNr and GPe monosynaptic connections to the PPN cholinergic and glutamatergic neurons using rabies tracing13–17. However, inputs to the GABAergic neurons have only been described in terms of the entire mesencephalic locomotor region (MLR), which extends beyond the PPN14. While SNr projections to MLR GABAergic neurons were identified but cannot be isolated to the PPN, GPe projections to the MLR GABAergic neurons were not detected14. Of note, these previous studies consider the PPN as a whole rather than separating its rostral and caudal subregions. While viral tracing can provide insight into neuroanatomical connections, limitations of these techniques19–21 and evidence that the rostral and caudal PPN may have distinct functions22–25 prompt a need for electrophysiological studies to determine whether the SNr and GPe inputs differentially innervate regionally-defined PPN neurons.

Optogenetic manipulations of the molecularly-defined PPN cell types have resulted in variable, and sometimes contradictory, locomotor and valence behaviors7,14,16,17,22–33. More recent studies have refined our understanding of these opposing behavioral effects by selectively stimulating the rostral or caudal GABAergic and glutamatergic PPN neurons22–25. In the basal ganglia, direct stimulation of molecularly-defined SNr and GPe subpopulations can also have distinct effects on locomotion and valence34–42. Despite the potential for the basal ganglia to modify behavior through the motor brainstem14, it is not clear how the SNr and GPe influence the rostral and caudal PPN to modify locomotion and valence.

Here, we use ex vivo whole-cell patch clamp electrophysiology paired with optogenetics to selectively stimulate the SNr or GPe axons while recording inhibitory inputs to the cholinergic, GABAergic, and glutamatergic PPN neurons to comprehensively characterize the strength and pattern of inhibition across the rostrocaudal PPN axis. We find that the GPe preferentially inhibits caudal GABAergic and glutamatergic PPN neurons whereas the SNr most strongly inhibits a caudal ‘hotspot’ of PPN glutamatergic neurons. Stimulating SNr or GPe axons over the PPN in vivo evokes opposing valence processing outcomes with place aversion during SNr stimulation and place preference during GPe stimulation. Surprisingly, and counter to the predictions of the canonical basal ganglia model, the GPe decreases locomotion through its projections to the PPN. While both the SNr and GPe inhibit the PPN, our results show that each nucleus differentially modulates the activity of specific cell types in the rostral and caudal PPN and are implicated in non-canonical basal ganglia circuits for modulating locomotion and valence processing.

Results

SNr and GPe axons display distinct distribution patterns across the rostral and caudal PPN

To characterize the axon distribution pattern from the SNr and GPe throughout the rostral and caudal extent of the PPN, we stereotaxically injected a channelrhodopsin virus with a fluorescent marker (AAV1-hSyn-ChR2-eYFP) into either the SNr or GPe of ChAT-Cre/Ai9-tdTomato mice (Figure 1A&C). We sliced 200 μm sagittal slices from mouse brains three weeks after viral injection, and cleared the brain slices using the CUBIC clearing method43. Using the red fluorescent ChAT-positive neurons to define the borders of the PPN, we found that SNr axons appear in both the rostral and caudal PPN. The SNr axons fill the rostral PPN evenly, but appear more distinctly clustered in specific areas of the caudal PPN (Figure 1B), whereas GPe axonal projections more densely fill the caudal region and strikingly avoid the rostral region (Figure 1D). These anatomical findings predict that the SNr will inhibit both rostral and caudal PPN neurons, while the GPe will selectively inhibit the caudal PPN. However, this axon imaging method does not reveal whether these axons form functional synapses with the neurons in the PPN and does not provide information on the strength of inhibition from each structure onto specific PPN cell types. Therefore, to evaluate the functional connectivity and synaptic characteristics from these inhibitory basal ganglia output structures to regionally- and molecularly-defined PPN neural subpopulations, we used ex vivo whole-cell electrophysiology to systematically record the inhibitory input to the rostral and caudal PPN from the SNr and GPe in three genetic mouse lines (ChAT-cre, Vgat-cre, and Vglut2-cre) which label cholinergic, GABAergic, and glutamatergic PPN subpopulations, respectively (Figure S1).

SNr and GPe axons display distinct distribution patterns across the rostral and caudal PPN.

(A,C) Stereotaxic injection of AAV1 delivering hSyn-ChR2-eYFP to the SNr or GPe of ChAT-Cre/Ai9-tdTomato mice, respectively. (B, D) Confocal images of EYFP-filled SNr or GPe axons across the PPN, respectively. CNII: cranial nerve II, scp: superior cerebellar peduncle; PB: parabrachial nucleus

The SNr inhibits cholinergic neurons in both the rostral and caudal PPN to a similar extent, but evokes stronger rebound in rostral neurons

To evaluate the functional connection between the SNr and cholinergic PPN neurons, we performed ex vivo whole-cell patch clamp electrophysiology paired with optogenetics to record synaptic currents in ChAT-Cre/Ai9-tdTomato mice injected with ChR2 in the SNr three weeks prior (Figure 2A). Neurons were targeted for whole-cell patch clamp in both the rostral and caudal PPN across three sagittal slices to evaluate the medial to lateral extent of the PPN (Figure 2B). To measure optically-evoked inhibitory post-synaptic currents (oIPSCs), cholinergic PPN neurons were held at −50 mV in whole-cell voltage clamp and blue (470 nm) light was applied (2 ms pulse duration, 13 mW) to activate the ChR2-infected SNr axons. In connected neurons, inhibitory currents were recorded in the presence of glutamatergic receptor blockers (AP5, NBQX, CNQX) during stimulation (Figure 2C). In a subset of neurons, a GABA-a receptor blocker, GABAzine, was applied to confirm the oIPSCs were GABA-mediated (Figure 2C). GABAzine eliminated SNr-evoked inhibitory currents in most cells, but a small residual current remained in some cells (see Methods for details). We found that all rostral (n=19/19 cells, N=6 mice) and essentially all caudal (n=26/27, N=6) recorded cholinergic PPN neurons responded to SNr axon activation with an observable oIPSC (Figure 2D).

SNr inhibition of rostral and caudal ChAT+ PPN neurons.

(A) Experimental set up to identify red ChAT+ PPN neurons for whole-cell patch clamp while stimulating ChR2-filled SNr axons [N=6]. (B) White arrowheads pointing to neurobiotin-filled patched neurons within the PPN across three 200µm slices. (C) Example trace of the first five oIPSCs [blue] in the 2-second 20 Hz train inhibited by GABA-a receptor blocker, GABAzine [green], while holding the cell at −50mV. (D) Percent connected among patched neurons in the rostral and caudal regions. (E) Average oIPSC amplitude at each of 40 optogenetic light pulses in n=15 rostral neurons and n=20 caudal neurons. (F) Left, Individual cell data for the first oIPSC amplitude and, right, example current traces. (G) Cell mapping of patched neuron locations with the first oIPSC amplitude represented by the color scale. (H) Normalized current amplitudes in E. (I) Left, Individual cell data for the PPR between the first two oIPSC amplitudes in the train and, right, example current traces. (J) Example voltage traces of action potential firing during a 2-second 20 Hz train stimulation in rostral (left) and caudal (right) neurons. (K) Percent of pre-optical stimulation firing frequency during stimulation and rebound in n=14 rostral vs n=23 caudal neurons. (L) Individual cell data for the absolute change in frequency during optical stimulation [ΔFrq During Opto]. (M) Individual cell data for the absolute change in rebound frequency post-stimulation [ΔRebFrq]; rostral vs. caudal p=0.0142. (N) Correlation analysis, color scale representing Spearman r [-1,1] and size representing p-value [1,0]. (O) Negative correlation between the absolute change in frequency during stimulation and post-stimulation rebound; r=-0.372, p=0.039. * p<0.05; bar graph data represent mean ± SEM; box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles.

To determine whether the SNr input to the rostral and caudal PPN cholinergic neurons differed in synaptic strength, we measured the amplitude of each oIPSC in a 20 Hz train of 2 ms optical stimulations applied for 2 seconds. There was no difference in amplitude of the first oIPSC in the stimulus train between rostral vs caudal cholinergic neurons (mean ± SEM; first oIPSC n=15 Rostral: 92.6 ± 18 pA, n=20 Caudal: 115.2 ± 17 pA; unpaired t-test, t=0.9041, df=33, p=0.3725; Figure 2E&F). Going forward, the first oIPSC amplitude of the train will be referred to as the first oIPSC.

While the rostral and caudal anatomical separation of PPN neurons is commonly used, it is a relatively coarse division. Therefore, we carefully mapped the location of each recorded cholinergic PPN neuron by matching slices to their bregma reference in the Paxinos and Franklin’s mouse brain atlas and aligning the midline of cholinergic neuron distribution which separates the loosely spread rostral neurons and densely packed caudal neurons (See Methods for details, Figure 2G). The strength of the SNr inhibitory connection is depicted by a color scale representative of the first oIPSC (Figure 2G, right). Throughout both the rostral and caudal PPN, cholinergic neurons receiving larger inputs from the SNr are intermixed with neurons receiving smaller inputs.

To compare short-term synaptic plasticity characteristics between rostral and caudal cholinergic PPN neurons, we normalized the amplitude of each current in the train to the first current. The oIPSC amplitude remained relatively constant with each subsequent pulse in both rostral and caudal cholinergic neurons (Figure 2H). In addition, the paired-pulse ratio (PPR) of the peak of the first two currents in the stimulation train did not differ (mean ± SEM; PPR n=15 Rostral: 0.96 ± 0.04, n=20 Caudal: 0.97 ± 0.05; unpaired t-test, t=0.1082, df=33, p=0.9145; Figure 2I). SNr synaptic inputs onto cholinergic neurons do not undergo significant short-term synaptic plasticity in either the rostral or caudal PPN.

Although oIPSC amplitude is indicative of the strength of a synapse, current size does not always correlate with functional impact on a neuron’s action potential output20,44,45. Because cholinergic PPN neurons display spontaneous firing46–50, we characterized the impact of SNr-mediated inhibition on action potential firing rate. While optically stimulating SNr axons with the same 20 Hz blue light protocol as above, we observed a decrease in firing rate in both rostral and caudal cholinergic PPN neurons Figure 2J-K). However, there was no significant difference between the percent frequency change in rostral and caudal neurons during SNr axon stimulation (mean ± SEM; %Frequency n=14 Rostral: During Stimulation 35.62 ± 8.87 %, Post Stimulation 136.1 ± 7.803 %; n=23 Caudal: During Stimulation 52.11 ± 8.43 %, Post Stimulation 119.1 ± 4.614 %; 2way ANOVA, StimulationF(1, 35)=81.9 p<0.0001, RegionF(1, 35)=0.002052 p=0.9641, InteractionF(1, 35)=3.273 p=0.0790; Figure 2K) or in the absolute decrease in firing frequency during optical stimulation [median (IQR); ΔFrq During Opto n=14 Rostral: −2.47 Hz (−3.37 to −1.44 Hz), n=23 Caudal: −2.10 Hz (−3.11 to −0.84 Hz); Mann Whitney, U=131, p=0.3602; Figure 2L]. These data show that the SNr equally inhibits action potential firing in rostral and caudal PPN cholinergic neurons.

Post-inhibitory rebound firing has been observed in the PPN as well as other brainstem structures49,51–57 and is associated with greater excitability, temporal encoding, and generation of oscillatory activity58. Indeed, we found that the rostral and caudal cholinergic PPN neurons exhibited an increase in action potential firing after inhibition (Figure 2K). This is in agreement with previous literature showing rebound spikes following inhibitory postsynaptic potentials recorded in PPN neurons of male rats during electrical stimulation of the SNr51,52. However, rostral cholinergic neurons had greater rebound activity compared with caudal cholinergic neurons as measured by the absolute change in rebound frequency [median (IQR); ΔRebFrq n=14 Rostral: 1.64 Hz (1.00 to 2.17 Hz), n=22 Caudal: 0.83 Hz (0.42 to 1.14 Hz); Mann Whitney, U=79, p=0.0142; Figure 2M]. This aligns with previous findings showing that rostral cholinergic neurons tend to have stronger intrinsic rebound activity compared to caudal neurons49. Importantly, we show that such rebound can be evoked by synaptic inhibition from the SNr.

To detect trends among SNr connectivity characteristics and intrinsic electrophysiological properties of cholinergic PPN neurons, we performed a correlation analysis to determine the strength and directionality of the relationship among five parameters [first oIPSC, PPR, ΔFrq During Opto, pre-optical stimulation frequency (Pre-Opto Frq), and ΔRebFrq; Figure 2N]. We found a weak but significant negative correlation between the frequency decrease during SNr axon stimulation and the rebound frequency increase (Spearman r=-0.372, p=0.039, Figure 2N&O). This finding is expected since strong inhibition is more effective at producing rebound activity than weak inhibition. However, rostral PPN cholinergic neurons do not receive significantly stronger inhibition from the SNr compared with caudal cholinergic neurons (Figure 2F&L). Therefore, larger rebound frequency in rostral neurons is likely due to differences in intrinsic cellular properties rather than synaptic properties49. Put together, these findings show that the SNr inhibits rostral and caudal cholinergic PPN neurons to a similar extent, but evokes stronger rebound firing in the rostral neurons.

The SNr inhibits GABAergic neurons in both the rostral and caudal PPN to a similar extent

While there are several rabies tracing studies suggesting SNr axons project to the cholinergic and glutamatergic PPN neurons13–17, SNr projections to the GABAergic PPN neurons are inferred from a study of the entire mesencephalic locomotor region14 and a characterization of their synaptic strength has not been conducted. Therefore, we evaluated the synaptic characteristics between the SNr and GABAergic neurons in the rostral and caudal PPN. In Vgat-Cre/Ai9-tdTomato mice, we injected ChR2 into the SNr to optically stimulate axons over the PPN while patching Vgat+ GABAergic PPN neurons (Figure 3A). We post-hoc stained the slices for ChAT+ neurons to verify the location of all patched GABAergic neurons within the cholinergic PPN. We found that all rostral (n=12/12, N=6) and most caudal (n=20/22, N=6) recorded GABAergic PPN neurons received inhibitory input from the SNr (Figure 3B).

SNr inhibition of rostral and caudal Vgat+ PPN neurons.

(A) Experimental set up to identify red Vgat+ PPN neurons for whole-cell patch clamp while stimulating ChR2-filled SNr axons [N=6]. (B) Percent connected among patched neurons in the rostral and caudal regions. (C) Average oIPSC amplitude at each of 40 optogenetic light pulses in n=9 rostral neurons and n=13 caudal neurons. (D) Left, Individual cell data for the first oIPSC amplitude and, right, example current traces. (E) Cell mapping of patched neuron locations with the first oIPSC amplitude represented by the color scale. (F) Normalized current amplitudes in C. (G) Left, Individual cell data for the PPR between the first two oIPSC amplitudes in the train and, right, example current traces. (H) Percent of pre-optical stimulation firing frequency [% Pre-Opto Frq] during stimulation and and rebound in n=7 rostral and n=19 caudal neurons. (I) Individual cell data for the absolute change in frequency during optical stimulation [ΔFrq During Opto]. (J) Correlation analysis, color scale representing Spearman r [-1,1] and size representing p-value [1,0]. (K) Negative correlation between the absolute change in frequency during stimulation and first oIPSC amplitude; r=-0.755, p=0.001. (L) Negative correlation between the pre-optical stimulation firing frequency and the PPR; r =-0.706, p=0.002. Box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles.

Although there was no major difference in the proportion of GABAergic neurons receiving inhibitory input from the SNr in the rostral vs caudal PPN, we compared the current amplitudes to characterize the synaptic strength and pattern of this connection. The oIPSC amplitudes across the 2-second stimulation as well as the first oIPSC showed no difference between rostral and caudal GABAergic neurons [median (IQR); first oIPSC n=9 Rostral: 76.5 pA (27.0 to 209.9 pA), n=13 Caudal: 59.7 pA (37.7 to 269.7 pA); Mann Whitney, U=50, p=0.6005; Figure 3C&D]. We also did not observe significant differences in short-term synaptic plasticity between rostral and caudal GABAergic neurons [median (IQR); PPR n=9 Rostral: 1.01 (0.88 to 1.16), n=13 Caudal: 1.06 (0.96 to 1.15); Mann Whitney, U=48, p=0.5123; Figure 3 F&G]. Defining short-term depression as a PPR<0.95 and short-term potentiation as PPR>1.05, we found 3/9 rostral and 7/13 caudal GABAergic PPN neurons showed short-term potentiation, while 3/9 rostral and 3/13 caudal GABAergic PPN neurons showed short-term depression. These data suggest additional heterogeneity within the GABAergic PPN subpopulation. Similar to the SNr connection with cholinergic PPN neurons, SNr input to the GABAergic PPN does not result in significant short-term synaptic plasticity and does not differ in amplitude or characteristics between rostral and caudal regions.

Because GABAergic PPN neurons display spontaneous firing50, we recorded action potential firing in current clamp while optically stimulating the SNr axons. The percent inhibition was not significantly different between rostral and caudal neurons (mean ± SEM; %Frequency n=7 Rostral: During Stimulation 37.91 ± 15.91 %, Post Stimulation 100.0 ± 14.57 %; n=19 Caudal: During Stimulation 30.68 ± 8.03 %, Post Stimulation 113.2 ± 5.187 %; 2way ANOVA, StimulationF(1, 24)=44.63 p<0.0001, RegionF(1, 24)=0.09642 p=0.7588, InteractionF(1, 24)=0.8841 p=0.3565; Figure 3H). There was also no difference in the absolute frequency change during optical stimulation between rostral and caudal neurons [median (IQR); ΔFrq During Opto n=7 Rostral: −5.03 Hz (−5.55 to −1.34 Hz), n=19 Caudal: −4.29 Hz (−11.02 to −2.94 Hz); Mann Whitney, U=55, p=0.5336; Figure 3I]. These data show that the SNr equally inhibits action potential firing in rostral and caudal GABAergic PPN neurons and does not evoke significant rebound firing in either neural population.

Overall, we found that the SNr inhibits the rostral and caudal GABAergic PPN neurons to a similar extent, but observed greater variability in synaptic strength and impact on neuronal action potential firing among caudal GABAergic neurons. One caudal subgroup responded with very small (<60 pA) amplitude currents and another subgroup responded with currents more than triple that amplitude, with one neuron responding with a 2.2 nA inhibitory current. We mapped the location of each recorded GABAergic neuron, as described above, and found no specific location corresponding with the neurons receiving stronger SNr inhibition across the PPN landscape (Figure 3E). This strong variability suggests that GABAergic PPN neurons may be divided into functionally distinct neural types that do not correspond to their rostrocaudal anatomical location. To determine if the neurons receiving larger inhibitory currents have distinct characteristics, we performed a correlation analysis (Figure 3J). As expected, we found that the absolute frequency reduction during inhibition is correlated with the first oIPSC amplitude recorded in a cell (Spearman r=-0.755, p=0.001; Figure 3K). Surprisingly, however, we found a significant negative correlation between PPR and the pre-optical stimulation frequency (Pre-Opto Frq, Figure 3L). Specifically, input to slower-firing neurons displayed greater short-term synaptic facilitation. This finding supports the idea that multiple functional neural subpopulations may be present within the GABAergic PPN population. Together, these results show that the SNr functionally inhibits GABAergic neurons across the rostrocaudal extent of the PPN, but suggests the presence of additional heterogeneity within the GABAergic PPN population.

The SNr differentially inhibits the rostral and caudal glutamatergic PPN neurons

Previous rabies tracing anatomical studies have identified SNr inputs to glutamatergic PPN neurons14,16,17, but a full characterization of their synaptic strength and rostral and caudal connectivity has not been conducted. In Vglut2-Cre/Ai9-tdTomato mice, we injected ChR2 into the SNr to optically stimulate axons over the PPN while patching Vglut2+ glutamatergic neurons in the PPN (Figure 4A). While holding cells at −50mV and applying 470 nm blue light stimulation, neurons were identified as connected if oIPSCs were observed. Similar to our findings in cholinergic and GABAergic neurons, we found that all rostral (n=19/19, N=6) and all caudal (n=28/28, N=6) recorded glutamatergic neurons received inhibitory input from the SNr (Figure 4B).

SNr inhibition of rostral and caudal Vglut2+ PPN neurons.

(A) Experimental set up to identify red Vglut2+ PPN neurons for whole-cell patch clamp while stimulating ChR2-filled SNr axons [N=6]. (B) Percent connected among patched neurons in the rostral and caudal regions. (C) Average oIPSC amplitude at each of 40 optogenetic light pulses in n=13 rostral neurons and n=13 caudal neurons. (D) Left, Individual cell data for the first oIPSC amplitude and, right, example current traces; p=0.0035. (E) Cell mapping of patched locations with the first oIPSC amplitude represented by the color scale. (F) Normalized current amplitudes in C. (G) Left, Individual cell data for the PPR between the first two oIPSC amplitudes in the train and, right, example current traces. (H) Example voltage traces of action potential firing during a 2-second 20 Hz train stimulation in rostral and caudal neurons, top to bottom. (I) Percent of pre-optical stimulation firing frequency [% Pre-Opto Frq] during stimulation and rebound in n=13 rostral and n=17 caudal neurons. (J) Individual cell data for the absolute change in frequency during optical stimulation [ΔFrq During Opto]; p=0.0197. (K) Spontaneous frequency in n=11 rostral and n=16 caudal neurons; p=0.0343. (L) Correlation analysis, color scale representing Spearman r [-1,1] and size representing p-value [1,0]. (M) Positive correlation between the absolute change in frequency during stimulation and PPR, r=0.486, p=0.030. (N) Negative correlation between the absolute change in frequency during stimulation and first oIPSC amplitude; r=-0.841, p<0.00001. (O) Negative correlation between the absolute change in frequency during stimulation and pre-optical stimulation frequency; r=-0.791, p<0.0001. (P) Positive correlation between the first oIPSC amplitude and pre-optical stimulation frequency; r=0.818, p<0.0001. * p<0.05, ** p<0.01; box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles.

Comparing the oIPSC amplitudes across the 20 Hz train, we discovered that the caudal glutamatergic neurons receive larger inhibitory currents than rostral glutamatergic neurons (Figure 4C). The median first oIPSC measured in caudal glutamatergic neurons was also significantly larger than those recorded in rostral neurons [median (IQR); first oIPSC n=13 Rostral: 61.6 pA (19.2 to 113.9 pA), n=13 Caudal: 313.5 pA (143.8 to 729.1 pA); Mann Whitney, U=29, p=0.0035; Figure 4D]. After post-hoc staining for the cholinergic PPN and mapping the location of each recorded glutamatergic PPN neuron as described above, we identified a ‘hotspot’ of strong SNr inhibition in a medial-caudal group of glutamatergic neurons (Figure 4E). This finding shows that a subset of caudal PPN neurons preferentially receive exceptionally strong SNr input.

To determine whether the SNr input to the rostral and caudal PPN show differential short-term synaptic plasticity, we compared the normalized oIPSC amplitudes. We found that inhibitory inputs to the rostral glutamatergic neurons display short-term synaptic facilitation with amplitudes increasing in subsequent stimulations (Figure 4F). However, strong facilitation occurred in only a few neurons that drove the mean upward and there was no significant difference between the PPR observed in rostral and caudal glutamatergic neurons when activating SNr axons [median (IQR); PPR n=13 Rostral: 1.07 (0.86 to 1.38), n=13 Caudal: 0.92 (0.89 to 1.02); Mann Whitney, U=77, p=0.7241; Figure 4G].

Because glutamatergic PPN neurons spontaneously fire16,50, we wanted to determine if larger inhibitory currents mediated larger decreases in the firing rate of caudal PPN neurons. Therefore, we recorded tonic action potential firing in glutamatergic PPN neurons while optically stimulating SNr axons (Figure 4H). Although inhibition measured as the percent of pre-optical stimulation firing frequency was not significantly different between rostral and caudal neurons (mean ± SEM; %Frequency n=13 Rostral: During Stimulation 32.62 ± 10.21 %, Post Stimulation 100.3 ± 6.113 %; n=17 Caudal: During Stimulation 24.71 ± 7.30 %, Post Stimulation 97.48 ± 2.228 %; 2way ANOVA, StimulationF(1, 28)=133.6 p<0.0001, RegionF(1, 28)=0.5214 p=0.4762, InteractionF(1, 28)=0.1765 p=0.6776; Figure 4I), we found that the absolute frequency decrease was significantly larger in caudal glutamatergic neurons compared to rostral neurons during SNr stimulation [median (IQR); ΔFrq During Opto n=13 Rostral: −3.21 Hz (−12.21 to −1.95 Hz), n=17 Caudal: −14.08 Hz (−20.04 to −6.07 Hz); Mann Whitney, U=55, p=0.0197; Figure 4J]. To examine the discrepancy between the percent decrease and the absolute decrease in firing frequency, we compared the spontaneous firing frequency between rostral and caudal glutamatergic neurons in cells that had no holding current applied. We found that caudal glutamatergic neurons fire faster than rostral glutamatergic neurons [median (IQR); Spontaneous Frequency n=11 Rostral: 6.8 Hz (5.4 to 17.2 Hz), n=16 Caudal: 20.6 Hz (11.2 to 37.2 Hz); Mann Whitney, U=45, p=0.0343; Figure 4K]. Together, these findings show that fast-firing caudal glutamatergic PPN neurons receive larger SNr-mediated IPSCs and display greater decreases in their action potential firing.

Since we found that the caudal glutamatergic neurons receive larger inhibitory currents and display greater decreases in firing rate during SNr axon stimulation, we performed a correlation analysis to evaluate the strength of the relationship among the synaptic and firing characteristics (Figure 4L). Interestingly, there was a weak but significant positive correlation between the PPR and frequency change during stimulation showing that input displaying short-term synaptic depression occurs in neurons with greater decreases in firing frequency (Spearman r=0.486, p=0.030; Figure 4M). As expected, SNr-mediated oIPSCs with larger amplitudes are correlated with greater decreases in firing frequency during stimulation (Spearman r=-0.841, p<0.00001; Figure 4N). Greater pre-optical stimulation firing frequency is correlated with larger frequency decreases during stimulation (Spearman r=-0.791, p<0.0001; Figure 4O). This finding is expected because the fast-firing neurons can be more strongly inhibited in terms of absolute decrease in action potential frequency (i.e. a floor effect). Surprisingly, however, pre-optical stimulation firing frequency is correlated with the first oIPSC amplitude (Spearman r=0.818, p<0.0001; Figure 4P), indicating that the SNr is more strongly connected to glutamatergic PPN neurons with faster firing rates.

These correlations highlight strong selectivity in SNr projections to PPN glutamatergic neurons. Overall, these findings support that the SNr selectively targets and strongly inhibits fast-firing caudal glutamatergic PPN neurons compared to rostral neurons, and that there is a medial-caudal hotspot where SNr inhibition of PPN glutamatergic neurons is particularly strong.

The SNr most strongly inhibits caudal glutamatergic PPN neurons

Most of the previous PPN circuit work has been done with rabies tracing. Aside from technical limitations, rabies tracing can only indicate the presence of synaptic connections. Our electrophysiological findings comprehensively characterize the strength and characteristics of the SNr input to each of the PPN cell types in both the rostral and caudal regions. Using this comprehensive dataset, we compared SNr input across the different PPN cell types in each region under the same electrophysiological conditions. We found that the median amplitude of the first SNr-mediated oIPSC was the same among rostral PPN neurons irrespective of cell type [median (IQR); first oIPSC n=15 ChAT+: 74.7 pA (44.5 to 136.5 pA), n=9 Vgat+: 76.5 pA (27.0 to 209.9 pA), n=13 Vglut2+: 61.6 pA (19.2 to 113.9 pA); Kruskal-Wallis, H(2,34)=0.7002 p=0.7046; Figure 5A].

The SNr most strongly inhibits caudal glutamatergic PPN neurons.

(A,C) Individual cell data for the first oIPSC amplitude recorded in each cell type for rostral and caudal PPN neurons, respectively. (B,D) Individual cell data for the absolute change in frequency during stimulation in each cell type for rostral and caudal PPN neurons, respectively. (E) Graphical depiction of SNr stimulation results. * p<0.05, ** p<0.01, **** p<0.0001; box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles.

The impact of these inhibitory inputs on neuronal firing did not differ with cell type as shown by the absolute change in frequency during SNr stimulation [median (IQR); ΔFrq During Opto n=14 ChAT+: −2.47 Hz (−3.37 to −1.44 Hz), n=7 Vgat+: −5.03 Hz (−5.54 to −1.34 Hz), n=13 Vglut2+: −3.21 Hz (−12.21 to −1.95 Hz); Kruskal-Wallis, H(2,31)=2.389 p=0.3028; Figure 5B].

We found that the SNr-mediated oIPSCs in caudal glutamatergic neurons were significantly larger than in cholinergic and GABAergic caudal PPN neurons [median (IQR); first oIPSC n=20 ChAT+: 109.3 pA (58.5 to 168.6 pA), n=12 Vgat+: 59.7 pA (37.7 to 269.7 pA), n=13 Vglut2+: 313.5 pA (143.8 to 729.1 pA); Kruskal-Wallis, H(2,43)=8.717 p=0.0128; Dunn’s test, ChAT+ vs Vgat+ p=0.6409, ChAT+ vs Vglut2+ p=0.0042, Vgat+ vs Vglut2+ p=0.0295; Figure 5C].

Likewise, the impact of SNr inhibition on neuron firing rate was strongest in caudal glutamatergic neurons. The absolute frequency decrease during stimulation in caudal glutamatergic neurons was larger than both cholinergic and GABAergic neurons [median (IQR); ΔFrq During Opto n=23 ChAT+: −2.10 Hz (−3.11 to −0.84 Hz), n=19 Vgat+: −4.29 Hz (−11.02 to −2.94 Hz), n=17 Vglut2+: −14.08 Hz (−20.04 to −6.07 Hz); Kruskal-Wallis, H(2,56)=24.99 p<0.0001; Dunn’s test, ChAT+ vs Vgat+ p=0.0049, ChAT+ vs Vglut2+ p<0.0001, Vgat+ vs Vglut2+ p=0.0332; Figure 5D]. While we found no significant difference between SNr-mediated oIPSCs recorded in GABAergic and cholinergic neurons (Figure 5C), the GABAergic neuron firing frequency was more strongly inhibited than cholinergic (Figure 5D). The greater reduction in firing frequency may be driven by the caudal GABAergic subgroup receiving particularly large oIPSCs (Figure 3D) and could be explained if GABAergic PPN neurons have higher input resistance than cholinergic neurons, as has been previously suggested59. Altogether, our findings show that the SNr most strongly inhibits the glutamatergic neurons followed by GABAergic neurons in the caudal PPN and more weakly inhibits rostral PPN neurons without any cell type bias (Figure 5E).

The GPe preferentially, but weakly, inhibits a subset of caudal GABAergic and glutamatergic PPN neurons

GPe projections to the cholinergic and glutamatergic PPN neurons have been shown in previous rabies tracing studies14–17, but projections to the GABAergic MLR neurons was not detected14. To evaluate the synaptic strength of these connections and determine regional connectivity, we repeated the previous whole-cell patch clamp experiments paired with optogenetics, but this time injecting ChR2 into the GPe of ChAT-Cre/Ai9-tdTomato, Vgat-Cre/Ai9-tdTomato, and Vglut2-Cre/tdTomato mice (Figure 6A). If oIPSCs were observed in response to optical stimulation, the neuron was identified as connected. In a subset of neurons, a GABA-a receptor blocker, GABAzine, was applied to confirm the oIPSCs were GABA-mediated (See Methods for details; Figure 6B). We recorded GPe-mediated oIPSCs in all three cell types of the PPN. As predicted by the low density of GPe axons present in the rostral PPN (Figure 1D), very few rostral PPN neurons received synaptic input from the GPe. This was true within each molecularly-defined PPN cell type [caudal vs rostral % connected; ChAT+: 30% vs 0% (n=9/30 vs 0/7), Vgat+: 68% vs 38% (n=21/31 vs 6/16), Vglut2+: 69% vs 26% (n=24/35 vs 5/19); Figure 6Ci-iii]. These findings show that the GPe preferentially targets caudal PPN neurons compared to rostral neurons. Since so few rostral PPN neurons received GPe inhibition, we focused our characterization of GPe inhibition to the caudal PPN.

GPe inhibition of the three PPN cell types.

(A) Experimental set up to identify red ChAT+, Vgat+, and Vglut2+ PPN neurons for whole-cell patch clamp while stimulating ChR2-filled GPe axons [N=6]. (B) Example trace of the first five oIPSCs in the 2-second 20 Hz train [blue] inhibited by GABA-a receptor blocker, GABAzine [green], while holding the cell at −50mV. (C) Left, Percent connected among patched neurons in the rostral and caudal regions and, right, cell mapping of patched locations with the first oIPSC amplitude represented by the color scale. Top to bottom, i. ChAT+, ii. Vgat+, and iii. Vglut2+ datasets. (D) Average oIPSC amplitude at each of 40 optogenetic light pulses in n=6 ChAT+, n=19 Vgat+, and n=15 Vglut2+ caudal PPN neurons. (E) Left, Individual cell data for the first oIPSC amplitude and, right, example current traces. (F) Normalized current amplitudes in C. (G) Left, Individual cell data for the PPR between the first two oIPSC amplitudes in the train; p=0.0206. Right, top, example current trace of short-term synaptic facilitation in VgAT+ neurons. Right, bottom, example current traces of short-term synaptic depression in Vgat+ and Vglut2+ neurons. (H) Percent of pre-optical stimulation firing frequency [%Pre-Opto Frq] during and post-stimulation in n=25 ChAT+, n=18 Vgat+, and n=29 Vglut2+ caudal PPN neurons. (I) Individual cell data for the absolute change in frequency during stimulation [ΔFrq During Opto].

(J) Correlation analysis for Vgat+ neurons, color scale representing Spearman r [-1,1] and size representing p-value [1,0]. (K) Negative correlation between the absolute change in frequency during stimulation and first oIPSC amplitude; r=-0.627, p=0.044. (L) Correlation analysis for Vglut2+ neurons.

(M) Negative correlation between the absolute change in frequency during stimulation and the pre-stimulation firing frequency; r=-0.648, p=0.014. *p<0.05, **p<0.01; box plots show median line with boxes showing IQR and whiskers showing 9th and 91st percentiles.

The percentage of connected caudal neurons was greater than the percentage of connected rostral neurons in both GABAergic and glutamatergic neurons (one-sided Fisher’s exact test; Vgat: OR=0.2857, CI=0.09190 to 0.9391, p=0.047; Vglut2: OR=0.1637, CI=0.05069 to 0.5767, p=0.0033; Figure 6Cii&iii), but we found no statistical difference between the rostral and caudal connectivity in cholinergic neurons (OR=0, CI=0 to 1.301, p=0.1150; Figure 6Ci). The number of connected caudal cholinergic neurons was not significantly greater than its 0% rostral connectivity (Figure 6Ci). In addition, the average oIPSC amplitude across the train and the first oIPSC measured in the connected cholinergic neurons was small [median (IQR); first oIPSC n=6 ChAT+: 22.6 pA (9.75 to 50.20 pA); Figure 6D&E]. Among recorded caudal cholinergic PPN neurons, both connected and non-connected, these small inhibitory currents displayed negligible effects on the neuronal firing frequency of the cholinergic PPN population in both the percent of pre-optical stimulation frequency and absolute frequency change [median (IQR); n=29 ChAT+: %Frequency During Stimulation 96.67 % (85.1 to 100.7 %); ΔFrq During Opto −0.16 Hz (−0.90 to 0.03 Hz); Figure 6H&I]. These findings indicate that the GPe has essentially no direct effect on the cholinergic PPN neural population. For this reason, we exclude GPe inhibition of cholinergic PPN neurons in the synaptic characterizations shown in Figures 6F&H. Together, we show that the GPe preferentially inhibits caudal GABAergic and glutamatergic neurons while largely avoiding rostral and cholinergic PPN neurons.

To determine whether the synaptic strength and characteristics from the GPe to the PPN differed between GABAergic and glutamatergic PPN subtypes, we evaluated the amplitude of the GPe-mediated oIPSCs during the 2-second 20 Hz optical stimulation. Both GABAergic and glutamatergic neurons responded to GPe stimulation with oIPSCs ranging from ones to hundreds of picoamperes (Figure 6D). While there was no significant difference in the first oIPSC measured in GABAergic and glutamatergic neurons [median (IQR); first oIPSC n=19 Vgat+: 24.77 pA (6.72 to 54.17 pA), n=15 Vglut2+: 65.71 pA (14.50 to 133.4 pA); Kruskal-Wallis, H(2,37)=3.050, p=0.2176; Figure 6E], the normalized amplitudes show that GPe input to GABAergic neurons facilitates with subsequent stimulations, while the input to glutamatergic neurons depresses (Figure 6F). The GPe-mediated inhibitory input to glutamatergic neurons displayed a significantly lower PPR than to GABAergic neurons [median (IQR); PPR n=19 Vgat+: 0.97 (0.76 to 1.2), n=15 Vglut2+: 0.81 (0.55 to 0.93); Mann-Whitney, U=76, p=0.0206; Figure 6G]. While short-term synaptic plasticity patterns were variable among GABAergic neurons with some depressing and others facilitating, GPe input to glutamatergic neurons generally depresses.

Short-term synaptic depression can dampen the impact that repeated IPSCs have on action potential output. Therefore, we compared the firing frequency change in each cell type during optogenetic stimulation of GPe axons. The PPN GABAergic neuron population firing rate was decreased to about 58% of their pre-stimulation firing with a median frequency decrease of 1.92 Hz during stimulation (Figure 6H&I). Similarly, the glutamatergic neuron population firing rate decreased to about 64% of their pre-stimulation firing with a median frequency decrease of 0.89 Hz during stimulation (Figure 6H&I). We found no statistical difference between GPe inhibition of firing frequency in caudal GABAergic and glutamatergic PPN neurons (mean ± SEM; %Frequency n=29 ChAT+: During Stimulation 92.39 ± 7.59 %, Post Stimulation 104.10 ± 5.23 %; n=18 Vgat+: During Stimulation 57.70 ± 10.11 %, Post Stimulation 91.81 ± 6.98 %; n=24 Vglut2+: During Stimulation 64.15 ± 7.63 %, Post Stimulation 104.7 ± 5.39 %; 2way ANOVA, StimulationF(1, 69)=25.46 p<0.0001, Cell typeF(1, 69)=4.444 p=0.0153, InteractionF(1, 69)=2.805 p=0.0674; Fisher’s LSD, df=138, %Frequency During Stimulation: ChAT+ vs. Vglut2+ p=0.0035, ChAT+ vs. Vgat+ p=0.0016, Vglut2+ vs. Vgat+ p=0.5378, %Frequency Post Stimulation: ChAT+ vs. Vglut2+ p=0.8883, ChAT+ vs. Vgat+ p=0.2838, Vglut2+ vs. Vgat+ p=0.2184; median (IQR); ΔFrq During Opto n=25 ChAT+: −0.16 Hz (−0.90 to 0.03 Hz), n=18 Vgat+: −1.92 Hz (−3.85 to −0.60 Hz), n=29 Vglut2+: −0.89 Hz (−3.28 to −0.43 Hz); Kruskall-Wallis, H(2,69)=17.21, p=0.0002; Dunn’s test, ChAT+ vs Vgat+ p=0.0002, ChAT+ vs Vglut2+ p=0.0009, Vgat vs Vglut2+ p=0.4020; Figure 6H&I). These findings show that the GPe inhibits GABAergic and glutamatergic neural populations to a similar extent.

To determine if there are significant relationships among the synaptic and firing characteristics, we performed a correlation analysis. Among GABAergic neurons receiving GPe inhibition, we found that the amplitude of the first oIPSC is significantly correlated with the absolute change in frequency during stimulation (Spearman r=-0.627, p=0.044, Figure 6K). Although GPe input to GABAergic neurons exhibited various PPRs, the correlation analyses revealed no significant relationship between PPR and absolute change in frequency during stimulation (Figure 6J). These findings suggest that the initial oIPSC amplitude of GPe input to GABAergic PPN neurons has a stronger relationship with firing rate than short-term synaptic plasticity does.

As nearly all GPe inputs to glutamatergic neurons exhibited short-term synaptic depression, we expected the correlation analysis to reveal a significant relationship between PPR and the change in firing frequency. However, similar to the GPe input to GABAergic neurons, PPR was not predictive of impact on neuronal activity (Figure 6L). GPe inhibition of the glutamatergic neurons showed a significant correlation between the absolute change in frequency during stimulation and pre-stimulation firing frequency (Spearman r=-0.648, p=0.014, Figure 6M). However, because the inhibitory synaptic current amplitude did not correlate with firing frequency, this correlation is likely due to a floor effect in which faster firing neurons have a greater capacity for reductions in firing frequency. Together, these findings show that the GPe preferentially inhibits caudal GABAergic and glutamatergic PPN neurons.

In vivo activation of GPe and SNr axons in the PPN show differential effects on locomotion and valence

Once we established that the SNr and GPe inhibit different profiles of regionally- and molecularly-defined subpopulations of the PPN using ex vivo optogenetic electrophysiology, we wanted to evaluate the behavioral consequences of selectively stimulating these inputs in vivo. Direct optogenetic activation of PPN subpopulations have been previously shown to either promote or inhibit motion, sometimes with contradictory results14,16,17,22–28. To determine whether differential inhibitory inputs onto the PPN influenced locomotor activity, we bilaterally injected wild-type mice with ChR2 in either the SNr or GPe and implanted an optical fiber over the PPN (Figure 7A&B).

In vivo activation of GPe and SNr axons in the PPN show differential effects on ocomotion and opposite effects on valence.

(A) Experimental set up to stimulate ChR2-filled SNr or GPe axons over the PPN in vivo. (B) Representative image of optical fiber tract overlaid with the pproximate optical fiber placement for SNr-[green] and GPe-[orange] injected mice. (C) Distance raveled over time in an open field with 1 minute 20 Hz 4.25 mW optical stimulations over the PPN in N=9 ontrol (Ctrl) mice (black circles), N=8 mice injected with ChR2 in the SNr (green diamonds), and N=9 mice injected with ChR2 in the GPe (orange hexagons); vertical blue lines represent periods of optical timulation. (D) Average distance traveled for each mouse across the six 1-minute optical stimulations in he high (4.25 mW) and low laser power (0.25 mW) conditions (E) Representative mouse track tracings uring real time place preference task in a three-chamber box and continuously stimulating EGFP- or ChR2-filled axons over the PPN at 20 Hz in SNr- and GPe-injected mice when the mice are in the timulation zone. (F) Percent time spent in the stimulation zone in N=16 control mice, N=9 mice injected with ChR2 in the SNr, and N=10 mice injected with ChR2 in the GPe with 4.25 or 0.25 mW laser power. G) Percent time spent in the stimulation zone during the first minute reintroduced to the RTPP box on ay 2 of RTPP with the laser off (no chamber is stimulated) Black line = median. * p<0.05, ** p<0.01, *** <0.001, **** p<0.0001; box plots show median line with boxes showing IQR and whiskers showing 9th nd 91st percentiles. See related supplemental Figures S2-S4 and Videos S1-S4.

Control mice were injected with an eGFP virus and implanted. After three weeks, we placed each mouse in the open field to measure gross locomotor behavior with and without optical stimulation of the SNr or GPe axons. We tracked the distance traveled for each mouse at baseline and during bilateral stimulation of either the SNr or GPe by applying blue light (473 nm, 4-4.5 mW) at 20 Hz with 2 ms pulses (4% duty cycle), for 1 minute at a time (Figure 7C). We found that stimulation of these two basal ganglia inputs to the PPN resulted in differential motor behaviors. Specifically, stimulation of SNr axons in the PPN increased distance traveled, while stimulation of GPe axons in the PPN decreased distance traveled [mean ± SEM; N=9 (3M, 6F) Control (Ctrl): 2.22 ± 0.24 m, N=8 (5M, 3F) SNr: 4.11 ± 0.51 m, N=9 (4M, 5F) GPe: 0.57 ± 0.18 m; Welch ANOVA p<0.0001; Dunnett’s test, Ctrl4.25mW vs SNr4.25mW p=0.0263, Ctrl4.25mW vs GPe4.25mW p=0.0002; Figure 7D & Figure S2, see Video S1&2].

The PPN has been implicated in reward processing and stimulation of the PPN can be reinforcing7,28,29,31,35,60. Therefore, we evaluated the effect of GPe or SNr axon stimulation in the PPN on valence in the real-time place preference (RTPP) task. In a three-chamber apparatus, the mouse could freely move between chambers for ten minutes. Optical stimulation of the SNr or GPe axons in the PPN was applied when the mouse entered the stimulated chamber and remained on at 20 Hz (4% duty cycle) until the mouse exited that chamber (Figure 7E). To reduce the motor effects of optical stimulation, this experiment was conducted with unilateral stimulation only. Interestingly, we again found opposite effects when stimulating the two basal ganglia inputs to the PPN. Mice avoided the stimulated chamber when the SNr axons were stimulated, but preferred the stimulated chamber when the GPe axons were stimulated (mean ± SEM; N=16 (6M, 10F) Ctrl: 40.76 ± 2.0 %, N=9 (5M, 4F) SNr: 23.99 ± 5.2 %, N=10 (5M, 5F) GPe: 76.11 ± 3.7 %; Welch ANOVA, p<0.0001; Dunnett’s test, Ctrl4.25mW vs SNr4.25mW p=0.0437, Ctrl4.25mW vs GPe4.25mW p<0.0001, Ctrl4.25mW vs SNr0.25mW p=0.0695, Ctrl4.25mW vs GPe0.25mW p=0.0467; Figure 7F, see Video S3&4). To further control for potential motion confounds, we recorded the location preference of the mice in the absence of optical stimulation the day after the stimulated RTPP assay (regardless of whether they received low or high power on day 1). The mice were placed in the central, neutral chamber and their location was tracked for 10 minutes. During the first minute of this assay, 8 out of 9 mice who had SNr optical stimulation on the first day continued avoiding the stimulated chamber while 9 out of 10 mice which had GPe optical stimulation continued to prefer the previously stimulated chamber compared to the median percent time spent in the stimulated chamber for mice in the control condition [median (IQR); N=16 (6M, 10F) Ctrl: 47% (14.79 to 69.24 %), N=9 (5M, 4F) SNr: 1.833% (0 to 41.75 %), N=10 (5M, 5F) GPe: 92.83% (58 to 99.21 %); Kruskall-Wallis, p=0.0008; Dunn’s test, Ctrl vs SNr p=0.1433, Ctrl vs GPe 0.0312; Figure 7G).

In addition to sending projections to the PPN, the GPe and SNr send axon collaterals to multiple other brain areas12,37,61,62. Therefore, it is possible that stimulating these axons in the PPN could cause antidromic stimulation of the GPe and SNr cell bodies, resulting in the inhibition of non-PPN brain regions. However, it has recently been shown that 0.25 mW laser power stimulation of axons can prevent antidromic activation of cell bodies42. Therefore, we repeated these optogenetic behavioral experiments using 0.25 mW laser power. We found that the locomotion effect of SNr stimulation over the PPN was no longer present during low power stimulation, but GPe stimulation still significantly decreased locomotion (mean ± SEM; distance traveled N=9 (3M, 6F) Ctrl: 2.22 ± 0.24 m, N=6 (3M, 3F) SNr: 2.24 ± 0.50 m, N=8 (4M, 4F) GPe: 1.09 ± 0.22 m; Welch ANOVA p<0.0001; Dunnett’s test, Ctrl4.25mW vs SNr0.25mW p>0.999, Ctrl4.25mW vs GPe0.25mW p=0.0131; Figure 7D). In the RTPP task, SNr axon stimulation continued to evoke place aversion at the trend level and GPe axon stimulation continued to significantly evoke place preference (mean ± SEM; % time in stimulation zone N=16 (6M, 10F) Ctrl: 40.76 ± 2.1 %, N=6 (3M, 3F) SNr: 27.72 ± 4.0 %, N=8 (4M, 4F) GPe: 68.41 ± 8.5 %; Welch ANOVA, p<0.0001; Dunnett’s test, Ctrl4.25mW vs SNr0.25mW p=0.0695, Ctrl4.25mW vs GPe0.25mW p=0.0467; Figure 7F). These findings show that decreased locomotion and place preference are mediated by GPe inhibition of the PPN while aversion is mediated by SNr inhibition of the PPN.

The increase in locomotion observed when stimulating the SNr axons in the PPN with high laser power is unexpected given that previous studies directly stimulating SNr neurons report decreases in movement34,36. Because injecting constitutive AAVs into the SNr can infect both GABAergic neurons and adjacent dopaminergic neurons of the substantia nigra pars compacta (SNc), we wanted to determine whether stimulating dopaminergic axons in the PPN enhanced locomotion and evoked aversion. To test this, we injected a cre-dependent ChR2 into the substantia nigra of DAT-cre mice (3M heterozygotes and 3F homozygotes) and ran open field and RTPP behavioral assays. We found that selective stimulation of dopaminergic axons in the PPN increased locomotion at high laser power (Figure S3A) and had no effect on locomotion at low power (Figure S3C). This finding suggests that dopaminergic neurons in the SNc send collateral projections to both the PPN and other locomotor-related structures (e.g. the striatum) such that stimulation of PPN-projecting SNc neurons increases locomotion.

The effect of dopaminergic axon stimulation did not differ from whole substantia nigra stimulation in either high (mean ± SEM; N=8 (5M, 3F) SNr4.25mW: 4.107 ± 0.51 m, N=6 (3M, 3F) DAT-Cre4.25mW: 3.778 ± 0.28 m; unpaired t-test, t=0.5159, df=12, p=0.6153; Figure S3B) or low laser power (mean ± SEM; N=6 (3M, 3F) SNr0.25mW: 2.238 ± 0.58 m, N=9 (3M, 3F) DAT-Cre0.25mW: 1.984 ± 0.30 m; unpaired t-test, t=0.4350, df=10, p=0.6728; Figure S3C). On the other hand, we found that stimulation of dopaminergic axons in the PPN did not cause place aversion, but rather induced real-time place preference at both high and low laser power (Figure S3D-F). Selectively stimulating dopaminergic axons induced place preference behavior significantly different from and opposite to whole substantia nigra stimulation at high (mean ± SEM; N=9 (5M, 4F) SNr4.25mW: 23.99 ± 5.18 %, N=6 (3M, S3F) DAT-Cre4.25mW: 70.83 ± 10.06 %; unpaired t-test, t=4.545, df=13, p=0.0005; Figure S3E) and low laser power (mean ± SEM; N=6 (3M, 3F) SNr0.25mW: 27.72 ± 4.0 %, N=6 (3M, 3F) DAT-Cre0.25mW: 50.82 ± 4.8 %; unpaired t-test, t=3.697, df=10, p=0.0041; Figure S3F). Together, these findings show that the the nigral increase in locomotion is not mediated through the PPN, and that dopaminergic input from the SNc to the PPN is rewarding while GABAergic input from the SNr to the PPN is aversive.

Discussion

In this study, we comprehensively characterize the synaptic strength and impact on neuron action potential firing in molecularly- and regionally-defined PPN subpopulations while stimulating the SNr or GPe using ex vivo electrophysiology and optogenetics. We also stimulated the SNr or GPe axons in the PPN in vivo which revealed that the SNr and GPe evoke differential locomotion and valence processing behaviors in non-canonical basal ganglia circuits. Together, these findings show that distinct and selective inhibition of PPN subpopulations by the SNr and GPe can alter behavioral output.

Region- and cell type-specific inputs to the PPN

The SNr and GPe have been previously shown to form monosynaptic synapses with PPN neurons using rabies viral tracing12–17. While SNr input to cholinergic and glutamatergic PPN neurons has been reproduced across different studies13–17, SNr inputs to GABAergic PPN neurons has only been inferred from studies of the entire mesencephalic locomotor region (MLR), consisting of both the PPN and cuneiform nucleus14. Several studies have used slice electrophysiology to show that the SNr inhibits PPN neurons generally32 or to compare SNr input to cholinergic and non-cholinergic PPN neurons48,51,52. However, a full characterization of the SNr input to regionally- and molecularly-defined PPN neurons in adult animals has been lacking. Using ex vivo whole-cell patch recordings, we were able to measure inhibitory currents in each of the PPN cell types during SNr stimulation showing that the SNr functionally inhibits each cell type within the rostral PPN to a similar extent while most strongly inhibiting a medial-caudal ‘hotspot’ of glutamatergic neurons.

While the GPe has been shown to project to the cholinergic and glutamatergic PPN neurons15–17, our electrophysiological results show that GPe inhibition of the cholinergic neurons is weak to non-existent and its inhibition of the glutamatergic PPN neurons is caudally biased. In contrast to a previous study showing that the GPe does not project to GABAergic MLR neurons14, we find that the GPe also preferentially inhibits a caudal subgroup of GABAergic PPN neurons. While the GPe selectively inhibits a caudal subpopulation of GABAergic and glutamatergic PPN neurons, its inhibition is much weaker than SNr inhibition of the PPN in terms of the proportion of connected PPN neurons, the inhibitory current amplitudes, and the inhibition of neuronal firing. Overall, our findings utilize a systematic approach to characterize the synaptic strength between these inhibitory basal ganglia nuclei and the PPN across its rostrocaudal axis for each cell type allowing us to compare SNr and GPe to PPN circuitry with regional-level granularity. These results also encourage future work involving the PPN to consider the individual influence of its rostral and caudal regions.

Noncanonical GPe motor circuits

We found that stimulation of GPe axons over the PPN decreased locomotion in both low (putatively local) and high (putatively generating antidromic activity) laser power stimulations. This behavioral outcome appears counter to the canonical model of basal ganglia movement pathways, in which GPe activation increases movement. Although there is strong support for this canonical model, recent in vivo recordings have found subpopulations of GPe neurons that show activity patterns counter to this model (neurons that increase activity during immobility)63,64. Similarly, a subset of GPe neurons paradoxically increase activity upon indirect pathway striatal projection neuron stimulation41. While one explanation for these heterogeneous responses could be local inhibition within the GPe40,41,65,66, another possibility is that distinct subpopulations within the GPe differentially modulate locomotor behavior39–41,61,67. Our findings support non-canonical basal ganglia motor pathways involving GPe neurons that project to the PPN.

Our characterization of region-specific inhibitory inputs to the PPN helps us understand how GPe inhibition of the PPN can mediate decreases in locomotion. Our electrophysiology experiments show that the GPe exerts minimal influence on the rostral PPN neurons which appear to decrease locomotion22,24,25 and preferentially inhibits caudal GABAergic and glutamatergic PPN neurons. Because stimulating the caudal GABAergic and glutamatergic PPN neurons increases locomotion23, GPe selective inhibition of these caudal neurons could explain our behavioral experiments showing that stimulation of GPe axons over the PPN decreases locomotion. Future studies are needed to determine if GPe inhibition of both cell types is necessary for mediating decreased locomotion. Aligned with our findings, selectively stimulating the Npas1+ or FoxP2+ subpopulations in the GPe decreases locomotion39,41. Some groups have found that direct stimulation of the PV+ GPe neurons increases locomotion39,62; however, one study shows that inhibition of PV+ GPe neurons can increase locomotion42. While we show that GPe projections to the PPN decrease locomotion, future studies are needed to determine which GPe subpopulations project to the PPN to modulate locomotion.

GABAergic vs dopaminergic effects of nigral projections to the PPN

While we found that stimulating nigral axons in the PPN with high laser power increased locomotion, we observed no locomotor effect when using low laser power stimulation. The simultaneous inhibition of caudal PPN neurons that promote movement23 and rostral PPN neurons that suppress movement22,24,25 may explain a net zero effect on locomotion during low power optical stimulation of SNr axons in the PPN. Indeed, low laser power stimulation of nigral axons in the most rostral implant location resulted in increased locomotion, while the most caudal implant location resulted in decreased locomotion (Figure S4). Interestingly, the enhanced locomotion observed during high laser power stimulation suggests that a subset of nigral neurons projecting to the PPN promote movement. However, the SNr canonically suppresses movement and the direct somatic stimulation of PV+ and GAD2+ SNr neurons that send axon collaterals to the PPN12,68 decreases gross locomotor movement34,68. Since injecting constitutive AAVs into the substantia nigra can infect both GABAergic neurons of the SNr and adjacent dopaminergic neurons of the SNc, it is possible that stimulation of dopaminergic activity may mask the impact of stimulating the GABAergic SNr alone.

Both the increased locomotion with high laser power and the lack of change with low laser power was recapitulated when we ran these experiments while selectively stimulating dopaminergic projections from the SNc to the PPN (Figure S3). Therefore, our findings suggest that activation of the PPN-projecting dopaminergic SNc neurons promotes locomotion, but not through direct actions at the PPN. This is consistent with previous literature showing that SNc dopaminergic neurons send axons to both the PPN and striatum69 and that direct stimulation of nigrostriatal dopaminergic neurons can increase movement initiation and vigor70. Therefore, antidromic stimulation of the SNc may evoke dopamine release into the striatum to increase movement in our high laser power experiments. Future studies are needed to evaluate the motor behavioral impact of the different SNr cell types when inhibiting the entire PPN versus the rostral or caudal PPN subregions.

By contrast, we find that both high and low power non-selective SNr axon stimulation in the PPN is aversive (Figure 7E-G), while both high and low power selective stimulation of dopaminergic SNc axons in the PPN is rewarding (Figure S3). These findings are particularly interesting because they indicate that the dopaminergic vs GABAergic inputs from the substantia nigra to the PPN serve opposite functions regarding valence signaling. Importantly, these findings indicate that GABAergic input from the SNr is highly aversive and that this aversive signaling occurs through the PPN. Interestingly, these findings also show that dopaminergic SNc neurons do not require antidromic stimulation to signal reward. This means that dopaminergic signaling within the PPN may serve to reinforce behavior. Future work is needed to parse out specific reward and aversion pathways involving neural subpopulations within the SNr, SNc, and PPN.

Differential PPN inhibition by the SNr and GPe mediate opposing valence processing outcomes

We found that stimulation of SNr axons over the PPN was aversive in real time place preference (RTPP), while stimulation of GPe axons over the PPN led to a preference for the stimulated chamber. Previous studies directly stimulating SNr and GPe subpopulations have implicated these structures in reward and aversion processing.

Inhibition of Vgat+ SNr neurons has been shown to increase place preference36 while its direct stimulation has no effect34,36. However, selective stimulation of the PV+ SNr neurons induces aversion in RTPP34. Although the GPe has been predominantly studied in the context of movement, recent work has begun to reveal a role for the GPe in valence processing42,71–73. One recent study shows that inhibition of PV+ GPe neurons induced aversion while inhibition of Npas1+ GPe neurons induced placed preference in RTPP42. Because activation of PV+ SNr neurons is aversive34 and inhibition of PV+ GPe neurons is aversive42, it is likely that our results are due to activation of the PV+ SNr and GPe axons in the PPN. While we show that activation of the PPN-projecting SNr and GPe subpopulations strongly affect valence processing, future studies are needed to determine whether these effects are specifically mediated by PV+ SNr and GPe projections to the PPN.

Neurons in the PPN have been implicated in reward and aversive-related behaviors7,28,29,31–33,35,60. The cholinergic PPN neurons are implicated in positive valence with place preference and increased lever pressing through their projections to the ventral tegmental area (VTA)28,29. Aligned with our findings, direct inhibition of the cholinergic PPN neurons evokes place aversion28. Of particular significance, we show that the SNr inhibits the cholinergic PPN neurons while the GPe axons avoid cholinergic PPN neurons. Therefore, it is possible that the difference between the SNr and GPe effects on valence is due to the difference in their inhibition of PPN cholinergic neurons.

Direct stimulation of the glutamatergic PPN neurons has also been implicated in positive valence with increased reinforcement behavior7. Both the cholinergic and glutamatergic PPN neurons directly influence dopamine release in the striatum and dopaminergic neuron activity in the SNc and VTA7,28,29,74–78. Specifically, the caudal PPN neurons are thought to innervate the medial part of the SNc and the VTA9,78, two major reward pathway hubs. Our data show that the SNr most strongly inhibits the caudal glutamatergic PPN neurons. Therefore, strong SNr inhibition of these neurons would remove excitatory drive from the reward-related dopaminergic neurons. This pathway could also contribute to the aversive effect of SNr axon stimulation in the PPN. Future experiments are needed to test whether direct inhibition of the caudal glutamatergic PPN neurons receiving SNr input is sufficient to evoke aversion.

The pathways underlying the GPe-mediated rewarding effect are less clear and may indicate functional heterogeneity among caudal PPN neurons. We find that the GPe only inhibits a subset of caudal GABAergic and glutamatergic PPN neurons. It is possible that the subset of GABAergic PPN neurons targeted by the GPe disinhibits local cholinergic and glutamatergic neurons to increase excitation of the rewarding dopaminergic neuron pathways. Another possibility is that the specific GPe-inhibited PPN population may have particularly aversive-properties. For example, PPN neurons that excite amygdala nuclei evoke place avoidance35,60. In this case, GPe inhibition of these neurons could be rewarding by removing excitation of amygdala nuclei involved in negative valence processing. To understand the PPN circuitry involved in GPe-mediated place preference, future experimental work must determine whether inhibiting GPe-targeted GABAergic or glutamatergic PPN subpopulations alone is sufficient to evoke place preference.

Conclusions

We have systematically characterized SNr and GPe inputs across the rostrocaudal axis of the PPN for each cell type – cholinergic, GABAergic, and glutamatergic. We show that the SNr inhibits nearly every PPN cell recorded with differential strength determined by cell type and anatomical biases while identifying a medial-caudal ‘hotspot’ of glutamatergic neurons most strongly inhibited by the SNr. In contrast, the GPe strikingly avoids the cholinergic PPN neurons and weakly, but selectively, inhibits a subpopulation of caudal GABAergic and glutamatergic neurons. We proposed that the differential inhibition of regionally-biased, mixed-cell type PPN subpopulations can alter behavioral outputs. By stimulating these inhibitory basal ganglia axons over the PPN, we show that the SNr evokes place aversion while the GPe evokes place preference and decreases locomotion. Together, our findings show that the SNr and GPe mediate opposing valence processing outcomes through the PPN and support a non-canonical basal ganglia motor pathway in which the GPe decreases locomotion through its projections to the PPN.